Endopiriform neurons projecting to ventral CA1 are a critical node for recognition memory

Version 1

Abstract

Abstract

The claustrum complex is viewed as fundamental for higher order cognition; however, the circuit organization and function of its neuroanatomical subregions are not well understood. We demonstrated that some of the key roles of the claustrum complex can be attributed to the connectivity and function of a small group of neurons in its ventral subregion, the endopiriform (EN). We identified a subpopulation of EN neurons by their projection to the ventral CA1 (ENvCA1-proj. neurons), embedded in recurrent circuits with other EN neurons and the piriform cortex. Although the ENvCA1-proj. neuron activity was biased toward novelty across stimulus categories, their chemogenetic inhibition selectively disrupted the memory-guided but not innate responses of mice to novelty. Based on our functional connectivity analysis, we suggest that ENvCA1-proj. neurons serve as an essential node for recognition memory through recurrent circuits mediating sustained attention to novelty, and through feed forward inhibition of distal vCA1 neurons shifting memory-guided behavior from familiarity to novelty.

Introduction

The claustrum complex, an evolutionarily conserved brain region found across mammalian species, reptiles, and birds, is hypothesized to serve as a node for establishing higher order cognitive functions by coordinating neuronal activity on a global scale (1–3). This view is based on its extensive connections with many cortical and subcortical areas (4, 5). Some of these functions include sensory perception and attention, which may affect memory processing, including working memory, associative memory, and recognition memory (6–8). Consistent with this view, abnormalities of the claustrum complex is found in major cognitive disorders, including Alzheimer’s disease, schizophrenia, and attention-deficit/hyperactivity disorders (9–11). However, microcircuitry and functional segregation of the individual constituents of the claustrum complex have remained unclear.

The cytoarchitecture and genetic expression broadly divide the claustrum complex into dorsal and ventral parts, with the dorsal part typically referred to as “claustrum” and ventral part having its own nomenclature: “endopiriform (EN)” in rodents (12, 13). Current evidence indicates that the claustrum forms reciprocal connections with dorsal cortices containing sensory, motor and association areas generating diverse physiological effects depending on the cortical targets and their activity patterns (4, 5). In contrast, EN primarily provides inputs to the piriform cortex and limbic systems (14, 15), suggesting its distinct circuit organization and role in bridging olfactory information processed by the piriform cortex and memory processed by the limbic systems (16, 17).

One of distinct targets of EN, the ventral CA1 (vCA1), is known to coordinate in a range of behaviors related to exploration and recognition memory, and these functions are thought to be regulated by its afferent inputs in a domain-specific manner (18–24). We, therefore, hypothesized that EN is a key node for establishment of recognition memory. To address this, we set out to characterize circuit and function of EN neurons defined by their projection to vCA1 (EN vCA1-proj. neurons) using genetic tools and mouse models of social and object recognition memory.

We found EN vCA1-proj. neurons innervated multiple components of the limbic system except amygdala and prefrontal cortex and produced potent feedforward inhibitory control over vCA1 pyramidal neurons. During recognition memory test, activity of EN vCA1-proj. neurons were condensed around conspecifics or objects where mice spent most time on. However, disruption of EN vCA1-proj. activity only impaired memory-guided exploration of novel stimuli without affecting innate exploration induced by novelty. These findings demonstrate that EN subserves some of the key functions required for recognition memory governed by specific limbic system.

Results

Endopiriform represents a major afferent of the ventral CA1

To assess the significance of EN as a vCA1 afferents, we injected a retrograde tracer into the vCA1 of mice and compared the number of the retrogradely labeled neurons in the EN and other brain regions (Fig. 1A, B, Methods). We found labeled neurons in multiple areas including the septum, entorhinal cortex, and basolateral amygdala (Fig. 1C-F and Fig. S1A) (25, 26). In addition to these well-known vCA1 afferents, prominent labeling was observed in the area defined as EN in Allen Brain Atlas (Fig. 1G, Methods). The number of presynaptic neurons in EN was ∼10 fold lower than entorhinal cortex (data not shown). However, the normalized count of presynaptic neurons in EN was significantly greater than basolateral amygdala, lateral septum, and medial septum (Fig. 1H).

Endopiriform projects to the ventral CA1

(A) Schematic diagram of a retrograde tracer injection performed. CAV-2-Cre or CTB647 was used in combination with Ai14 or wild-type mice, respectively. (B) An example fluorescent image of injection site. (C) Distribution of retrogradely labeled vCA1-projecting neurons in entire brain as a result of injection in panel B. (D-G) Fluorescent images showing retrogradely labeled vCA1-projecitng neurons in septum (SP), entorhinal cortex (ENT), basolateral amygdala (BLA), and endopiriform (EN). Str: Striatum, CLA: Claustrum. (H) Box plots showing relative number of vCA1-projecting neurons in different brain areas. Cell number in each brain area were normalized by summed cell number of EN, BLA, SP (lateral), and SP (medial). Median value for EN, BLA, SP (lateral), and SP (medial) is 0.45, 0.24, 0.08, and 0.21, respectively. Statistics: EN vs BLA, SP (lateral), or SP (medial), p < 0.001 for all; BLA vs SP (lateral) or SP (medial): p = 0.002 and 0.694, respectively; SP (lateral) vs SP (medial) p = 0.023. One-way ANOVA, repeated comparison with Tukey-Kramer test. N = 5. (I) An example fluorescent image of horizontal section showing retrogradely labeled vCA1-projecting neurons in EN. A: anterior, P: posterior. (J) Top: Example map showing distribution of vCA1-projecting neurons in entire EN in one mouse. Bottom left: A plot showing a quantified distribution at bin size of 1 mm. Each bin was normalized to total number of labeled neurons in EN. Data from each mouse is shown in grey. Median is indicated in black. Bottom right: Box plots comparing a number of vCA1-projecting neurons in anterior and posterior EN. Data were normalized to a total number of labeled neurons in EN. Median value for anterior vs posterior was 0.39 vs 0.54. Statistics: p = 0.008, Rank-sum.

The vCA1-projecting neurons in the EN were spread along antero-posterior axis with two peak densities, one in anterior and another in posterior to the bregma (Fig. 1I, J). Number of labeled neurons was greater in posterior EN (Fig. 1J). In contrast, the same analysis along dorso-ventral axis showed a single dominant peak at the depth consistent with the location of EN (Fig. S1B). To further confirm the location of these vCA1-projecting neurons, we injected a retrograde tracer in one color into vCA1 and a retrograde tracer in another color into a cortical area known to receive projections from EN or “claustrum” representing a dorsal part of claustrum complex (Fig. 2A). We targeted the prefrontal cortex to label EN, and the anterior cingulate cortex, the motor cortex, and the dorsomedial entorhinal cortex to label claustrum (Fig. 2C-F) (27–30).

Axon branching of vCA1-projecting EN neurons

(A) Schematic diagram of injection performed and two hypothesized outcomes. Yellow color indicates double-labeled neurons. PFC: Prefrontal cortex, ACC: Anterior cingulate cortex, MC: Motor cortex, ENT(dm): Dorsomedial entorhinal cortex. (B-F) Example fluorescent images of injection site for each region indicated. (G-J) Example fluorescent images of retrogradely labeled neurons in CLA complex. vCA1-projecting neurons were colored in red and cortex-projecting neurons were colored in green. (K) Box plots showing percentage of double-labeled vCA1-projecting neurons for each cortical injection. Medial value for PFC, ACC, MC, and ENT(dm) was 28.6, 4.6, 0, and 1.2, respectively. Statistics: PFC vs MC or ENT(dm), p = 0.015 and 0.047, respectively. One-way ANOVA with post-hoc Kruskal-Wallis test. Slice and animal number: 16 and 4 for PFC, 20 and 5 for ACC, 10 and 3 for MC, and 18 and 4 for ENT(dm). (L) Schematic diagram of projection pattern found in the experiment shown in A. (M) Schematic diagram of the intersectional approach performed to label vCA1-projecting EN neurons. (N-O) Example fluorescent images showing vCA1-projectig axons in different brain areas. All images were from the same mouse. (P) Magnified images showing vCA1-projecting axons in cortical layers. Magnified areas are indicated in O and fig. S2A-C as dashed squares. PIR: Piriform cortex

We found there is a major spatial overlap in labeling of vCA1-projecting neurons with prefrontal cortex-projecting neurons, but not with other cortical projection neurons (Fig. 2G-J). Moreover, around 30% of vCA1-projecting neurons were found to be double-labelled when a cortical injection was made into the prefrontal cortex, suggesting some vCA1-projecting neurons send axons to this cortical area (Fig. 2K, L).

To further determine the cortical and subcortical innervation pattern of vCA1-projecting EN neurons, we used an intersectional approach to label their somata and axons with eGFP (Fig. 2M). Apart from vCA1, axons from vCA1-projecting neurons were found in various cortical areas including prefrontal cortex, lateral entorhinal cortex, and piriform cortex (Fig. 2N-P and Fig. S2A-C). Projection to prefrontal cortex was sparse relative to other areas, as expected based on the retrograde labeling data (Fig. 2K). The connectivity of vCA1-projecting EN neurons to the amygdala, which represents the major component of limbic systems, was also sparse relative to vCA1 (Fig. S2C), indicating vCA1 is the main target of these EN neurons in this system.

Taken together, these findings indicate EN represents a major afferent of vCA1. Moreover, EN neurons projecting their axons to vCA1 also send strong collaterals to lateral entorhinal cortex and piriform cortex, but relatively weak to the prefrontal cortex or other limbic structures. Since our subsequent study focused on EN neurons defined by their projection to vCA1, we refer to them as ENvCA1-proj. neurons.

Projection pattern and synaptic connectivity of EN axons in vCA1

We next investigated projection pattern of EN axons in hippocampus by injecting an anterograde tracer (AAV-GFP) into the EN (Fig. 3A, B). Analysis of different parts of hippocampal section indicated EN axons were mostly confined in stratum lacunosum-moleculare (SLM) layer of distal part of vCA1 (Fig. 3C, D). Projection to intermediate or dorsal CA1 was limited or undetectable (Fig. S3A-C).

EN projection pattern in vCA1

(A) Schematic diagram of injection performed. (B) An example fluorescent image of horizontal brain section containing injection site in EN. (C) An example fluorescent image of ventral hippocampal slice with EN axons resulted from the injection in (B). Distal and proximal vCA1 where GFP signals were measured were indicated by dashed lines. (D) Normalized GFP signals across the layer. All signals were normalized by maximum GFP signal in the distal subregion. Median of area under the curve: distal; 11.12, proximal; 5.99. distal vs proximal, p = 0.016. Wilcoxon signed-rank test. n = 8 slices (N = 4 mice). SR: stratum radiatum, Py: pyramidal layer.

Since SLM mainly consists of GABAergic neurons (31), we hypothesized that EN axons form synapses with this cell type. To test this, we first applied monosynaptic rabies tracing technique to Vgat-Cre mice (Fig. 4A, Methods). The distribution of starter cells (identified by GFP and mCherry co-labeling) was confirmed to be within the ventral-intermediate region of hippocampus using AMaSiNe (Fig. 4B, S4A) (32). In these mice, the presynaptic neurons were consistently observed in EN in addition to other expected areas (e.g., septum and entorhinal cortex) (Fig. 4C and d, S4B-F) (33, 34). The density of the presynaptic neurons in EN was greater at the posterior region to the bregma (Fig. 4D), consistent with our earlier observation (Fig. 1J).

ENvCA1-proj. neurons target vCA1 interneurons

(A) Schematic diagram of injection performed for monosynaptic rabies tracing. (B) Location of starter cells in the hippocampus. Data from 5 brains were overlaid. (C) Example images showing presynaptic cells in EN (white arrow). (D) Top: Location of presynaptic EN neurons. Data from 5 brains were overlaid. Bottom: Plots indicate distribution of presynaptic cells in EN along AP axis. Cell counts in a given bin (1 mm) were normalized by total cell number in EN. Thin gray lines indicate individual data. A thick black line indicates median. Inset showing plot comparing cell number in anterior and posterior location as in Fig. 1J. Median: anterior vs posterior, 0.25 vs 0.72, p = 0.031; Wilcoxon signed-rank test. (E) Top: Schematic diagram of injection performed. Bottom: An example fluorescent image of injection site in EN. (F) Schematic diagram of recordings performed. Recordings were performed from identified GABAergic neurons sequentially in a random order. (G) An example confocal image of an acute vCA1 slice used for recordings. Recorded GABAergic neurons were filled with biocytin and post-labeled with Alexa488. White dashed lines indicate border of Py-SR and SR-SLM. (H) Median traces of photo-evoked EPSCs recorded from neurons at different laminar position. (I) Connection probability (CP) of EN axons and recorded neurons at different laminar positions. Numbers in bars indicate number of response positive neurons /total recorded neurons. (J) Left: Peak amplitudes of evoked EPSCs (normalized to total input from recorded slice). Right: the same as left but for mean amplitude. Gray circles indicate each neuron. Black lines indicate median. Median (peak amp): Py-SR vs SR vs SR-SLM vs SLM: 1.1 vs 1.0 vs 1.1 vs 0.0. [Py-SR,SR, SR/SLM vs SLM], p = 0.076, 0.111, 0.04; one-way ANOVA, repeated comparison with Tukey-Kramer test. Median (mean amp): Py-SR vs SR vs SR-SLM vs SLM: 1.0 vs 0.8 vs 1.4 vs 0.0. [Py-Sr, SR, SR-SLM vs SLM], p = 0.14, 0.251, and 0.114; one-way ANOVA, repeated comparison with Tukey-Kramer test. N = 5 mice.

To determine how EN inputs are integrated into the local circuit in vCA1, we systematically assessed the connections between EN axons and GABAergic neurons in different layers of distal vCA1. We expressed channelrhodopsin-2 (ChR2) into the EN axons by injecting AAV-ChR2-mCherry into EN and performed whole-cell recording from GABAergic neurons in acute vCA1 slices in the presence of TTX (1 µM) and 4-AP (100 µM) to isolate monosynaptic connection (35). To ensure the recording from GABAergic neurons, we used Vgat::Ai14 mice in which all GABAergic neurons are labeled with tdTomato (Fig. 4E-G). We recorded from 3 to 4 GABAergic neurons located in different laminar position in distal vCA1 (Py-SR border, SR, SR-SLM border, or SLM) in the same slice to construct a “laminar profile” of EN◊vCA1 inputs (Fig. 4F and G). Whole-field LED stimulation evoked excitatory postsynaptic current (EPSC) in most GABAergic neurons recorded. However, contrary to the expectation, connection probability and strength were lowest in neurons in SLM compared to neurons in other layers (Fig. 4H-J). Taken together, these results indicate EN axons preferentially innervate the GABAergic neurons spanning across the Py-SR border to SR-SLM border in distal vCA1.

EN axons produce feedforward inhibition onto vCA1 pyramidal neurons

SLM also contains the apical tuft dendrite of pyramidal neurons, which could be a target of EN axons in addition to GABAergic neurons. To test this, we expressed ChR2 in EN axons but made whole-cell recording from pyramidal neurons in distal vCA1 in the presence of TTX and 4-AP. Since CA1 pyramidal neurons can be differentiated into superficial and deep layer neurons with distinct anatomy and function (36), we recorded from neurons in each layer as well as nearby GABAergic neurons at the Py-SR and SR-SLM borders for the comparison (Fig. 5A). We found monosynaptic input to superficial and deep layer pyramidal neurons was similar, and monosynaptic input to pyramidal neurons (superficial and deep pooled) and GABAergic neurons was also comparable (Fig. S5).

EN disynaptically inhibits vCA1 pyramidal neurons

(A) A confocal image of recorded pyramidal neurons in vCA1. Recorded pyramidal neurons were filled with biocytin and post-labelled with Alexa488. Ori.: oriens, Super.: superficial layer pyramidal neuron. Deep: deep layer pyramidal neuron. (B) Example traces of photo-evoked EPSCs and IPSCs in superficial (black trace) and deep (gray trace) pyramidal neurons. Red trace: after TTX and 4AP. Blue trace: after gabazine. (C) Scatter plots of peak amplitudes of EPSCs and IPSCs for each neuron. Dashed line indicates a unitary line. Superficial neurons: empty Black circles. Deep neurons: Gray circles. N = 4 mice. (D) A summary diagram of EN◊vCA1 circuit. In vCA1, EN axons innervate GABAergic neurons (IN), which in turn inhibits pyramidal neurons (both superficial and deep).

Since EN axons innervated multiple GABAergic neurons across different layers, we hypothesized they may exert stronger inhibition onto pyramidal neurons through feed-forward inhibition. To test this, we recorded EPSC and inhibitory postsynaptic current (IPSC) from superficial and deep pyramidal neurons using cesium-based internal solution in the brain slices expressing ChR2 in the EN axons (Methods). At the command potential of −70 mV, photo-stimulation of EN axons evoked EPSCs in majority of pyramidal neurons in both layers (7 out of 10 neurons in superficial and 5 out of 10 neurons in deep) (Fig. 5B). At the command potential of +12 mV, the same stimulation evoked outward postsynaptic current in the same neurons, including those without EPSCs (10 out of 10 for superficial and deep) (Fig. 5B). This outward current was abolished by application of TTX and 4-AP or gabazine (10 µM), indicating it is disynaptically driven and mediated by inhibitory GABAA receptor (Fig. 5B). When relative strength of EPSCs and IPSCs were compared for each neuron, the strength of IPSCs overwhelmed the EPSCs in all cases (Fig. 5C). This indicates EN axons disynaptically inhibit pyramidal neurons in vCA1 (Fig. 5D).

ENvCA1-proj. neurons receive inputs from the piriform cortex

We next examined afferents that may drive the EN◊vCA1 circuit. To address this, retrograde CAV-2-Cre was injected into vCA1 to express Cre into ENvCA1-proj. neurons, then applied monosynaptic rabies tracing technique to specifically label its presynaptic neurons in wild type mice (Fig. 6A). Starter cells and presynaptic neurons were spatially mapped and quantified using AMaSiNe (Fig. 6B and C, S6). Since presynaptic neurons were sparse in contralateral hemisphere, data from ipsilateral hemisphere was analyzed. Moreover, we applied minimum labeling threshold to determine the brain region for further analysis (Methods). We found the piriform cortex contained highest number of presynaptic neurons followed by EN (Fig. 6D-F). Within piriform cortex, layer 2 neurons were the major source of afferent (Fig. 6G). These data indicate ENvCA1-proj. neurons receive major input from piriform cortex and form a recurrent connection with neurons in EN (Fig. 6H).

ENvCA1-proj. neurons receive inputs from the piriform cortex and within EN

(A) Schematic diagram of performed. (B) Location of starter cells in EN. Data from 6 brains were overlaid. (C) An example map of presynaptic cells in one brain. Top: Horizontal view. Injections were made in left hemisphere. Bottom: Front view. (D) Number of presynaptic cells in different brain areas (normalized to total presynaptic cell number). Median for AIA, AON, PIR, ENT(lateral), CLA, EN, Pallidum, and SI (in fraction): 0.7, 0.3, 3.8, 0.3, 0.5, 1.7, 0.3, and 0.2. [AON, ENT(lateral), CLA, Pallidum, SI] vs PIR: p = 0.016, 0.008, 0.048, 0.006, and 0.003. EN vs SI, p = 0.043. one-way ANOVA with post-hoc Kruskal-Wallis test. Full list of statistical tests and p values are in Table1. (E) An example fluorescent image of coronal slice containing EN and PIR and labeled presynaptic cells. (F) Example maps of presynaptic cells in PIR (top) and EN (bottom). (G) Number of presynaptic cells in PIR in different layers (normalized to the total number of labeled neurons in PIR). Median: L1 vs L2 vs L3, 0.11 vs 0.61 vs 0.27. L1 vs L2, p < 0.001; L1 vs L3, p = 0.005; L2 vs L3, p < 0.001; one-way ANOVA, repeated comparison with Tukey-Kramer test. (H) Summary diagram showing input-output circuit of ENvCA1-proj. neurons (red circle).

The activity of ENvCA1-proj. neurons is correlated with the time mice spend in space

The downstream target of EN, vCA1, is implicated in social, odor and object recognition memory (20–22). To determine whether and how the activity of EN neurons is related to this function, we expressed GCaMP8s in ENvCA1-proj. neurons and monitored their activity using fiber photometry while video recording mice performing social recognition memory test consisting of pretest, sociability test, and discrimination test (Fig. 7A, B, Methods). The nose position of the subject mouse was tracked during tests using DeepLabCut (Fig. 7C) (37, 38), and cumulative time and calcium event were mapped onto the arena space (Fig 7C, S7). We found these values were correlated in space across all sessions (i.e., the more time the mice spent in a given space, the more the ENvCA1-proj. activity), including in the open field with no object or in the test arena with conspecifics or objects (Fig. S7, S8A-B).

Activity of ENvCA1-proj. neurons predicts social/object discrimination performance

(A) Right and Middle: Schematic diagrams of injection and optic cannula implantation performed. Left: An example image showing jGCaMP8s expressing ENvCA1-proj. neurons and a tract of optic cannula. (B) The diagram of social recognition memory test. A test consisted of pretest (5 min), sociability test (10 min), and discrimination test (5 min). (C) Left: An example image of a mouse with nose point marker tracked by DLC. Right: An example map of nose points and calcium events in the arena during sociability test. (D) Nose distance vector (distance between subject’s nose point and a center point of novel/familiar mouse chambers) and simultaneously acquired calcium signals during discrimination test. Gray dashed lines indicate the border of interaction zones. Open red circles indicate the calcium event detected (threshold: Z > 2.58). A part of calcium signal trace was expanded at the bottom for clarity. (E) Left: Correlation between calcium event ratios (calcium event number in the mouse interaction zone/ calcium event number in the object interaction zone) and sociability index. p = 0.29, N = 8 mice. Right: same as in left panel, but for calcium event ratios (calcium event number in the unfamiliar mouse interaction zone/ calcium event number in the familiar mouse interaction zone) and discrimination index. p = 0.003, N = 8. (F) The diagram of novel object recognition memory test. A test consisted of habituation (5 min), familiarization (10 min), and discrimination test (5 min). (G) Same as in (E), but for calcium event number ratios (calcium event number in the unfamiliar object interaction zone/calcium event number in the familiar object interaction zone and discrimination index. p = 0.193. N = 5 mice.

Cumulative calcium events of ENvCA1-proj. neurons became high towards the space around objects or conspecifics when they are present and was correlated with interaction times except for sociability session (Fig. S8C-D). In the sociability session, the cumulative calcium events around a novel conspecific and object were similar despite mice spending more time on conspecifics (Fig. 7E left, S8C). In contrast, in the social discrimination test, both the cumulative time and cumulative ENvCA1-proj. activity were higher for an unfamiliar than familiar conspecific (Fig. 7D-E, S8C). As a result, ENvCA1-proj. activity and behavior had a greater degree of correlation in social discrimination than sociability.

We also tested if a similar correlation occurs in a non-social context by recording ENvCA1-proj. activity and behavior in the object discrimination test (Fig. 7F). The cumulative time and calcium events were significantly correlated across different sessions i.e., familiarization, and object discrimination (Fig. 7G, S8E), largely consistent with the findings in the social discrimination test.

These data indicate that the activity of ENvCA1-proj. neurons is generally correlated with the time the mouse spends in a particular location at the basal level (i.e. open field) or task conditions. However, the degree of correlation appears to vary depending on the session, such that it was highest in the pretest/familiarization and social/object discrimination test, but lowest in the sociability test.

Inhibition of ENvCA1-proj. neurons impairs social/object recognition memory

Although ENvCA1-proj. activity predicted the behavior in pretest/familiarization and discrimination test better than sociability test, their causality is unclear. To address this, we inhibited the activity of ENvCA1-proj. neurons using inhibitory DREADD (hM4Di) during social and object discrimination tests (Fig. 8A-C and h, S9A) (39). To specifically express hM4Di in ENvCA1-proj. neurons, CAV-2-Cre and AAV-flox-hM4Di were bilaterally injected into vCA1 and EN, respectively (Fig. 8A-B). hM4Di was replaced with tdTomato for control group. We made a protocol that allows within-subject comparison, such that a mouse goes through a test with chlozapine-N-oxide (CNO, 1 mg/ kg) treatment then the same test with saline treatment the next day, or vice versa (Fig. 8C and H). In social discrimination test, CNO treatment did not affect the pretest or sociability in hM4Di or control group, but specifically impaired social discrimination in hM4Di group (Fig. 8D-G, S9B-C). A similar effect was also observed in novel object discrimination (Fig. 8I, S9D).

Chemogenetic inhibition of ENvCA1-proj. neurons specifically impairs recognition memory

(A) Schematic diagram of injection performed. (B) Left: An example fluorescent image showing expression of hM4Di in ENCA1-proj. neurons. Right: Expansion of areas marked by white rectangle in image on the left. (C) The diagram of social recognition memory test. A session consisted of habituation (5 min), Pretest (5 min), Sociability test (10 min), and discrimination test (5 min). The group treated with a saline in session 1 was treated with CNO (1mg/ kg) in session 2 the following day. (D) Example cumulative times heat map in arena during sociability tests for the same mouse under two different treatments. (E) Plots comparing sociability index for hM4Di and tdTomato mice under two different treatments. Gray empty circles indicate each mouse. Gray thin lines show paired data from one mouse. Thick black bars indicate median. Median sociability index: hM4Di mice (Saline vs CNO), 0.75 vs 0.76, p = 0.391; signed-rank test; tdTomato mice (Saline vs CNO), 0.77 vs 0.75, p = 0.383; signed-rank test. (F) Example cumulative time heat maps in arena during discrimination tests for the same mouse under two different treatments. (G) Comparison of discrimination index for hM4Di and tdTomato mice under two different treatments. Gray empty circles indicate each mouse. Gray thin lines show paired data from one mouse. Thick black bars indicate median. Median discrimination index: hM4Di mice (Saline vs CNO), 0.66 vs 0.57, p = 0.042; signed-rank test; tdTomato mice (Saline vs CNO), 0.65 vs 0.63, p = 0.742; signed-rank test. (H) The diagram of novel object recognition memory test. A session consisted of habituation (5 min), familiarization (10 min), and discrimination test (5 min). CNO treatment was the same as in C. (I) Same as in G, but for novel object discrimination index. Median discrimination index: hM4Di mice (Saline vs CNO), 0.64 vs 0.5, p = 0.003; signed-rank test.

Apart from recognition memory, vCA1 is implicated in anxiety and associative fear memory (40, 41), thus, we tested contribution of ENvCA1-proj. neurons to these functions. The anxiety level as determined by locomotor activity and sociability was not affected by CNO treatment in hM4Di group (Fig. 8E, S9E). Similarly, fear memory as assessed by the freezing response during training (with trace fear conditioning) or recall to context or tone was not affected by CNO treatment in hM4Di group (Fig. S10).

Taken together, these data indicate that ENvCA1-proj. activity is not causally related to innate exploration behavior induced by novelty, or anxiety or fear memory. However, their activity is necessary for mice to discriminate between familiar and unfamiliar conspecific or object, suggesting their major role in general recognition memory.

Discussion

We characterized the circuit and function of EN neurons targeting specific subregions and layers of vCA1. In contrast to other parts of claustrum complex, ENvCA1-proj. neurons were reciprocally connected with the piriform cortex representing a major downstream target of the olfactory bulb (42). This circuit motif may predict the predominant role of ENvCA1-proj. neurons in social recognition memory, given the ability of social odor to engage multiple olfactory pathways innervating the piriform cortex (43). However, we found that these neurons contributed to broader domains of recognition memory. This is supported by the observations that the inhibition of ENvCA1-proj. activity impaired both social and non-social discrimination.

In all phases of recognition memory test, EN vCA1-proj. activity was highly correlated with the time mice spent in given locations and this correlation was biased towards novel stimuli when mice were engaged in their exploration irrespective of the stimuli being social or non-social (Fig. 9A). Taken together, these data suggest that the function of ENvCA1-proj. neurons was primarily related to the detection of and attention to novel stimuli. The recurrent connections within EN, which is another prominent feature of ENvCA1-proj. neurons’ network, potentially support these processes to help establish the recognition memory (44).

The role of ENvCA1-proj. neurons in recognition memory

(A) Summary of findings from in vivo experiments. (B) The model of ENvCA1-proj. circuit function in recognition memory. During memory-guided exploration, recurrent circuits in ENvCA1-proj. neurons maintain the attentional response to novel stimuli while regulating the degree of response to familiar stimuli in vCA1 via feedforward inhibition.

Attentional processes mediated by ENvCA1-proj. neurons can be considered of particular relevance for memory-guided behavior, given that inhibition of ENvCA1-proj. activity selectively impaired conspecific/object discrimination but not innate exploratory behavior provoked by novelty as seen in pretest, sociability, and familiarization (Fig. 9A). This functional specialization, likely associated with its unique circuit connectivity to the limbic system, could differentiate EN from claustrum.

Studies on EN are generally scarce, and the available evidence of EN projections to vCA1 suggests that these projections are sparse (14). Potential reasons include the difficulty in clearly delineating the EN from claustrum complex and non-uniform distribution of ENvCA1-proj. neurons along antero-posterior axis. By leveraging contemporary circuit analysis tools, we demonstrated that the EN→vCA1 circuit is non-trivial; ENvCA1-proj. neurons were more numerous than commonly studied vCA1 afferents, including those from the septum and amygdala. Moreover, these axons exhibited a potent disynaptic inhibition of distal vCA1 pyramidal neurons, indicating their major functional implications for vCA1 activity.

Our data indicated that although EN axons terminated at the SLM of distal vCA1, they exhibited stronger connections to GABAergic neurons spanning from the Py-SR border to the SR-SLM border than to SLM neurons. The detection of EN inputs from GABAergic neurons at the Py-SR border was surprising, considering the distance between soma position and the EN axons. However, there are several GABAergic cell types below SLM that extend their primary dendrites into SLM (45, 46), likely explaining the observed connectivity pattern. Among these are chandelier cells found at the Py-SR border, which preferentially inhibits the axon initial segments of pyramidal neurons, and are thus well suited to exert the powerful feedforward inhibition of vCA1 pyramidal neurons observed in our recordings (47).

There are several potential circuit effects by which EN inputs could affect vCA1 function. One is the induction of non-linear events in vCA1 pyramidal neurons by acting in concert with other afferents (e.g., lateral entorhinal cortex targeting SLM) (34) to generate a specific output pattern in pyramidal assemblies representing task-relevant stimuli. Alternatively, EN inputs may improve signal to noise ratio of the information conveyed through the lateral entorhinal cortex, CA2 (in the case of social context), and possibly other afferents (e.g., CA3) (48). Another scenario is tuning of vCA1 pyramidal neurons’ response to familiar stimuli through dynamic combination of a weak monosynaptic excitatory input and strong disynaptic inhibitory input. Such mechanism suggests that at a milliseconds scale novelty (EN) and familiarity (vCA1) recognition might occur though alternating rather than simultaneous activity of brain microcircuits. Taken together, we propose a model for the role of ENvCA1-proj. neurons in recognition memory by balancing memory-guided attentional responses to familiarity and novelty through combination of feedforward inhibition of vCA1 (a node for recognition of familiarity) and the recurrent circuits that contribute to sustain attention to novelty (Fig. 9B).

The model of promoting novelty detection by suppressing familiarity responses is consistent with previous observations showing that vCA1 pyramidal neurons predominantly respond to familiar stimuli whereas novelty response can only occur through activation of vCA1 interneurons (49, 50). Nevertheless, alternating activation of EN and vCA1 could prove essential for the behavioral relevance of EN activity associated with recognition memory, without affecting innate exploratory behavior to novelty. Accordingly, the latter behavior is primarily associated with the prefrontal cortex and amygdala, areas that are only weakly targeted by ENvCA1-proj. neurons (Fig. 2P, S2C) (51–53).

In addition to advancing our understanding of the basic organization and function of brain circuits underlying higher cognitive processes, our findings suggest dysfunction of ENvCA1-proj. neurons could be a key contributing factor to the deficits in claustrum complex function and recognition memory found in neuropsychiatric disorders (9, 10). Discrete EN populations can thus emerge as pathophysiological substrate, but also as important treatment targets of key symptoms of these disorders.

Materials and Methods

Mice

All experiments were performed in accordance with standard ethical guidelines and were approved by the Danish national animal experiment committee (License number: 2021-15-0201-00801). Unless otherwise noted, all mice used were C57BL6/J mice or transgenic mice with C57BL6/J backgrounds. They were 2-4 months old at the time of the experiment. Transgenic mice used were B6.Cg-Gt (ROSA)26Sortm14(CAG-tdTomato)Hze/J mice (Ai14, JAX007914) and B6J.129S6(FVB)-Slc32a1tm2(cre)Lowl/MwarJ mice (Vgat-Cre, JAX028862). Similar number of male and female mice were used for all experiments except for behavior tests. For behavior tests, male mice were used.

Viruses and retrograde tracers

The information on viruses used in our experiments was the following. CAV-2-Cre (PVM); AAV1-mDlx-HBB-chl-dlox-TVA_2A_oG(rev)-dlox (v271-1, VVF); AAV1-hSyn1-dlox-TVA_2A_mCherry_2A_oG(rev)_dlox (v306-1, VVF); Rabies-GFP (NTNU viral core facility); Rabies-mCherry (NTNU viral core facility); AAV5-CAG-GFP (37525-AAV5, Addgene); AAV5-CAG.hChR2(H134R)-mCherry (10054-AAV5, Addgene); AAV8-hSyn-DIO-hM4Di (Gi) – mCherry (44362-AAV8, Addgene); AAV5-hEF1a-dlox-EGFP(rev)-dlox (v217-5, VVF); and retrograde AAVrg-CAG-GFP (37825-AAVrg, Addgene), AAV1-hSyn1-dlox-jGCaMP8s(rev) (v627-1, VVF). We also used retrograde traces of red Retrobeads (Lumafluor) and Cholera toxin subunit B conjugated with Alexa 647 (CTB647, Thermo Fisher).

Stereotaxic injections

Stereotaxic injections were performed using a stereotaxic frame (Model 940, Kopf). Mice were anesthetized with isoflurane and were subcutaneously injected with buprenorphine (0.1 mg/kg) and metacam (1.5 mg/Kg) for post-operative pain relief. After incising the scalp over the cranium, a small hole was bored with a microdrill bit, and a beveled glass pipette (Wiretrol II, 5-000-2010, Drummond Scientific Company) back-filled with mineral oil and front-filled with the material to be injected was slowly inserted into a target coordinate. After injecting a small volume (50 - 100 nL) with a custom-made displacement injector (based on MO-10, Narishige), the pipette was left in place for 3 to 5 mins before slow retraction. The incision was closed with a nylon suture. The stereotaxic coordinates used were (relative to bregma, in mm): anteroposterior (AP) −3.0; mediolateral (ML) 3.12; dorsoventral (DV) 3.7 for ventral CA1; AP +2.0; ML 0.2; DV 1.5 and 2.5 for prefrontal cortex; AP 0.0; ML 0.2; DV 1.0 and 1.5 for anterior cingulate cortex; AP 0.0; ML 1.5; DV 0.3 and 0.7 for motor cortex; and AP −5.0; ML 3.0; DV 1.5 for dorsomedial entorhinal cortex. For endopiriform injection, we injected two sites from the following coordinates: AP −0.27, ML 3.2, DV 4.2; AP 0.0, ML 3.0, DV 4.2; or AP +1.0, ML 2.7, DV 4.0. Mice were thermally supported with a feedback-controlled heating pad maintained at ∼37 °C (ThermoStar Homeothermic system, RWD). Mice were used for experiments 3-5 weeks post injection.

Implantation of the optic probe

After stereotaxic injection of CAV-2-Cre into the vCA1 and AAV1-hSyn1-dlox-jGCaMP8s into the EN (AP -−0.27, ML 3.2, DV 4.2), a fiber optic cannula (400 μm core diameter, 4.5 mm length, NA 0.39, R-FOC-BF-400C-39NA, RWD) was inserted towards EN. Once in the correct depth, exposed brain surface around the implant was covered by Kwik-Cast (WPI) and was then fixed on skull with dental adhesive resin cement (Super-Bond, SUN MEDICAL).

Ex vivo electrophysiology

Mice were deeply anesthetized with isoflurane and decapitated. Horizontal sections (300 µm) containing ventral hippocampus were prepared by vibratome (VT1200S, Leica) in ice-cold choline solutions containing (in mM): 25 NaHCO3, 1.25 NaH2PO4-H2O, 2.5 KCl, 0.5 CaCl2, 7 MgCl2, 25 D-glucose, 110 Choline chloride, 11.6 Ascorbic acid, and 3.1 C3H3NaO3. Slices were subsequently incubated in artificial cerebrospinal fluid (ACSF) containing (in mM): 125 NaCl, 25 NaHCO3, 1.25 NaH2PO4-H2O, 2.5 KCl, 11 D-glucose, 2 CaCl2, and 1 MgCl at 34 °C for 30 mins then at room temperature (∼20 °C) for at least 1 hour before the recording.

Whole-cell recording was performed with an upright microscope (BX51WI, Olympus) equipped with a motor-controlled stage and focus (MP285A and MPC-200, Sutter instrument), differential interference contrast, coolLED (model pE300), and a monochrome camera (Moment, Teledyne). Neurons were visualized with a 60x lens (1.00 NA, LUMPlanFL N, Olympus) with the software Micro-Manager-2.0 gamma (US National Institutes of Health). Pipettes (4-5 MΩ) were pulled from thick-walled borosilicate capillary glass with a puller (model P-1000, Sutter instrument). For a voltage-clamp recording, the internal solution contained (in mM): 135 CsMeSO3, 10 HEPES, 4 Mg-ATP, 0.3 Na-GTP, 8 Na2-Phosphocreatine, 3.3 QX-314 (pH was adjusted to 7.35 with CsOH). In some experiments, a biocytin (4 mg/mL) and Alexa488/586 (50 μM) were further added for morphological studies. Recordings and hardware control were performed with Wavesurfer (Janelia Farm). Signals were amplified and Bessel filtered at 4 kHz with an amplifier (MultiClamp 700B, Molecular Device), then sampled at 10 kHz with a data acquisition board (USB-6343, National Instrument). Recording with access resistance change of > 20% from the baseline (∼30 MΩ or less) were discarded. Liquid junction potential was not corrected. All recordings were performed at ∼34 °C maintained with the inline heating system (TC-324C, Warner instrument).

To stimulated channelrhodopsin-2, a blue LED was delivered at a short pulse (5-msec) through a 4x objective lens (0.16 NA, UPlanSApo, Olympus). The light power (measured at the focal plane) was 20 mW for the recordings in normal ACSF and 40 mW for the recordings in the presence of tetrodotoxin and 4-aminopyridine. Photo-stimulation was repeated 3 to 6 times (at intervals of 10 or 20 sec) to obtain an average response trace. Photo-evoked responses less than 2 standard deviations of the baseline were considered as a “zero” response. The mean amplitudes indicate mean amplitudes between LED stimulus-onset to 50 ms of average responses.

Quantification of presynaptic neurons

Mice were deeply anaesthetized with a ketamine (120 mg/kg) and xylazine (24 mg/kg) mixture and were transcardially perfused with 4% paraformaldehyde (PFA). The brains were extracted and immersed in 4% PFA overnight followed by 20 and 30% sucrose solution for a cryoprotection. The brains were subsequently embedded in O.C.T compound (Tissue TEK, Sakura) in a mold and frozen on the dry ice. Coronal slices (100 µm) of entire brain were cut with Cryostat (CM3050S, Leica) and were washed with phosphate-buffered saline (PBS) before DAPI staining. All sections were mounted onto glass slides with a coverslip (thickness:0.13-0.17 mm, Hounisen) and mounting media (DAKO, S3023, Agilent Technologies).

The injection sites and its specificity were verified by registering the images into Allen Brain atlas using QuickNII and VisuAlign (54).

For quantification of presynaptic neurons and visualization of their distribution in 3D mouse brain, epifluorescence images acquired with Slidescanner (4231 x 3462 pixels, 10x objective, Olympus VS120) were analyzed with Matlab-based program AMaSiNe (32). For Fig.1, the coronal section between AP + 3 to −4 mm were registered in AMaSiNe and analyzed.

For the monosynaptic rabies tracing study in Fig. 6, the brain areas containing over 1% presynaptic cells of total presynaptic cells were proceeded further analysis. Fraction of presynaptic cells in the different areas were calculated by dividing a number of presynaptic cells in each brain areas by the average number of presynaptic neurons in all brain areas.

For the study in Fig. 2, the sections were immunostained with NeuN before imaging. Here, sections were immersed with 5% normal goat serum and 0.2% Triton X for 30 mins at room temperature before incubation with NeuN primary antibody (anti-rabbit, 1:1000) (ABN78, Milipore) at 4 °C for overnight. The sections were then washed with PBS and were further incubated with secondary antibody (goat anti-rabbit conjugated with Alexa 488 or 568, 1:500) (A11008 or A11036, Thermo Fisher) for 1 hour at room temperature before mounting. A single plane image of claustrum complex was acquired with confocal microscope (10x objective, 1024 x 1024 pixels, Zeiss). 2-4 images, range of AP: +1.4 to −0.2, were collected per brain. Fraction of double-labelled neurons (with different retrograde tracers) were calculated for each brain by dividing a total number of double-labeled neurons by the total number of vCA1-projecting neurons (labeled with one color).

Quantification of fluorescence signal from axons

For Fig. 3 and S3, horizontal section (100 µm) images were taken with Sliderscanner (10x lens, pixel: 2560 x 3072). All images from the same brain were taken with the same light intensity and detection setting. After background subtraction, GFP signals in ventral, intermediate, and dorsal CA1 column were normalized by the highest signal in ventral CA1 in the same brain.

For Fig. 2, coronal sections (100 µm) were immunostained with GFP antibody before imaging. Here, sections were washed with PBS, the sections were incubated in blocking solution containing 5% normal goat serum and 0.2% Triton X for 30 mins at room temperature before incubation with GFP primary antibody (anti-rabbit, 1:1000) (NB600-308, MNovus Biologicals) in blocking solution at 4 °C for overnight. The sections were then washed with PBS and were further incubated with secondary antibody (goat anti-rabbit conjugated with Alexa 488, 1:300) (A11008, Thermo Fisher) for 1 hour at room temperature before mounting. Images were taken with Slidescanner (10x, pixel: 3072 x 3584). Signal per pixel in ROI was calculated for each projection site after background subtraction.

Social memory test with chemogenetic inhibition of ENvCA1-proj. neurons

Social memory test is based on previous articles with slight modifications (55), (56). All tested mice were habituated to the handling for 3 days. Prior to the test day, we divided subject mice into two groups. On the 1st test day (session 1), one group received chlozapine-N-oxide (CNO, 1 mg/kg, i.p.) and another group received a vehicle. After 30 mins, the subject mice were placed in a three-chamber box for 5 mins (habituation), then the same box with two pencil chambers containing the same object in opposite corners for 5 mins (pretest). After 5 mins break in their home cage, the subject mice were again placed in the same box, but this time, an object in one pencil chamber was replaced with different object and an object in another pencil chamber was replaced with a conspecific (3-4 weeks old male, habituated to the handling and pencil chamber for 3 days prior to the session 1). This “sociability test” lasted for 10 mins and was followed by social discrimination test after 30 mins break in the home cage. The social discrimination test lasted for 5 mins and consisted of the same setting as the sociability test except an object in a pencil chamber, which was replaced with a new conspecific (i.e., a novel mice). In the following day, CNO and saline treatments were swapped between the groups, and all mice went through the same series of tests (session 2) for the within-subject comparison of the CNO effect. The position of the pencil chamber including an object or a conspecific were counterbalanced. The tests were performed under 6-10 lx room light and video was recorded at 25 frames/sec with GigE camera (Basler ace acA1300-60gm, Basler) placed at the ceiling.

Trace fear conditioning

After one week of social memory test, subject mice were examined for their tone-associated fear memory and extinction. One group received CNO while another group received vehicle 45 mins before the test. In conditioning day, mice were placed in a box with a metal grid floor and ethanol odor (context A) and after 3 mins were delivered with 20-sec pure tone (2 kHz, 20 dB) followed by 18-sec time gap (trace) and 2-sec footshock (0.5 mA). The exposure to the same tone-trace-shock sequence was repeated 2 more times with inter-trial interval of 2 mins. In the following day, mice were re-exposed to the context A for 3 mins to assess their fear memory to the context. From day 3, mice were placed in a new context, the context B (smooth floor, breach odor), and went through the same protocol as in conditioning day but without shocks. The process was repeated over 5 days to assess the extinction of the fear memory to the tone and trace. All test was performed in dark (but infrared light was on), and behavior was video recorded at 25 frames/sec with infrared GigE camera (Basler ace acA1300-60gm, Basler).

Novel object recognition test

The novel object recognition test is based on the previous article with slight modifications (55). The subject mice were divided into two groups and CNO treatment was done as same as social memory test. 40 mins after CNO/vehicle i.p. injection, the subject mice were placed in an open field box (50 x 50 cm) for 5 mins (habituation). In following familiarization session, two same shape objects are placed in the open field box, and then the subject mice explored for 10 mins. After 1 hour break in their home cage, the subject mice were again placed in the same box, but one object is replaced by a novel object. The subject mice are allowed to explore the box for 5 min in discrimination session.

Social memory test and novel object recognition test with in vivo fiber photometry

Optic probe implanted mice were habituated handling for 3 days prior the test. In test day, the subject mouse attached with patch code were placed in an open field box (50 x 50 cm) for 5 min (habituation). After habituation, the mouse was place in a home cage for resting and waiting following session. For pretest, pencil chambers containing the same objects were placed at the corners of the open field box. Pretest was tested for 5 min. For sociability test, a novel mouse (4 weeks old, male) or a novel object were placed in pencil chambers. Sociability test took 10 min. After sociability test, the subject mouse return to a home cage and rest for 20 min. After 20 min rest, the subject mouse was tested discrimination session for 5 min. In discrimination test, a novel mouse (4 weeks old, male) or a familiar mouse that was used as the stimulus mouse in sociability test were placed in pencil chambers. Both in sociability and discrimination test, positions of pencil chambers were counterbalanced between subject mice. The tests were performed under 6-10 lux room light and video was recorded at 60 frames/sec with GigE camera (Basler ace acA1300-60gm, Basler) placed at the ceiling. Video acquisitions were controlled by Bonsai. Simultaneous start of video recording and calcium signal recording was controlled by Raspberry Pi.

After one week of social memory test, subject mice were examined for novel object recognition test. The procedure, the arenas and objects were used the same as in the novel object recognition test with chemogenetic inhibition of ENvCA1-proj. neurons.

Behavior data acquisition and analysis

Ethovision (version 16 or 17, Noldus) was used for the hardware control and data acquisition and initial quantification for the behavior. Behavioral variables, including movement trajectory, times spent in predefined areas, total distance traveled, travelling velocity, and freezing were extracted and further analyzed/nested with custom functions in the Matlab (R2020b). For social memory test, interaction zones were defined as an area between the edge of pencil chamber and 3 cm away from the edge. When the nose of the subject mouse was in the zone, the time was counted as interaction time. Sociability index was calculated by dividing total interaction time with mouse by total interaction time with mouse and object. The discrimination index was calculated by dividing total interaction time with a novel mouse by total interaction time with a familiar mouse and a novel mouse. Data from mice were excluded if discrimination index was less than 0.5 during saline treatment. For trace fear conditioning test, mice were considered “freezing” if pixel change between frames was less than 0.1 % for more than 1 sec.

Deeplabcut (ver.2.3) was used for tracking of annotated body points of the subject mice connected with patch code for calcium imaging of ENvCA1-proj. neurons during social memory test and novel object recognition test. Nose points were used for further analysis using Matlab (R2020b). To define interaction zone, reference points was set at the center of pencil chamber. The areas that distance between nose point and reference point was less than 100 pixel for social memory test and 50 pixel for novel object recognition test were dedicated to interaction zone.

Fiber photometry and analysis

Population activity of ENvCA1-proj. neurons were recorded using fiber photometry (Doric Lenses). GCaMP8s was excited with 470 nm light at 20-30 μW measured at the tip (isosbestic point used was 415 nm (15 μW)). Calcium-dependent and independent signals were collected using lock-in mode. Signals were detected at 10x gain. Recorded signals were butterworth filtered (low-pass at 40Hz) and calcium signals (in z-score) were extracted using Doric neuroscience studio V6 (Doric Lenses). Signals were further down sampled to 60 Hz using Matlab (R2020b) signal analyzer to match the behavioral sampling data (60 fps). Z-score > 2.56 (alpha = 0.01) was determined to be a calcium event. For heat map of cumulative time and calcium events, spatial binning was 50 pixels. To make correlation map (Fig. S7 and S8A), correlations between column of cumulative time heat map and cumulative calcium event heat map were tested using Matlab corr function.

Statistical analysis

Data and statistical analysis were performed with MATLAB R2020b (Mathworks). Unless otherwise noted, group data was presented as median, and first and third quartiles are shown for dispersion. Statistics were performed using a non-parametric test (signed-rank test for paired data and rank-sum test for unpaired data). One-way ANOVA (post hoc test: Tukey-Kramer test or Kruskal-Wallis test) were applied for multi-group comparison. Two-way ANOVA were performed for comparison of multi-factors. Pearson correlation coefficient was applied for correlation analysis. Significance was defined as *p < 0.05, **p < 0.01, ***p < 0.001. Outlier was detected by Grubbe test (GraphPad, significant level: 0.05). Sample number were indicated as “n”, animal number were indicated as “N”.

Supporting information

eLife Assessment

This important study offers insights into the function and connectivity patterns of a relatively unknown afferent input from the endopiriform to the CA1 subfield of the ventral hippocampus, suggesting a neural mechanism that suppresses the processing of familiar stimuli in favor of detecting novelty. The strength of evidence is solid, with careful anatomical and electrophysiological circuit characterization, although the functional role of this pathway in behavior is not firmly established. The work will be of broad interest to researchers studying the neural circuitry of behavior.

Reviewer #1 (Public Review):

Summary:

The anatomical connectivity of the claustrum and the role of its output projections has, thus far, not been studied in detail. The aim of this study was to map the outputs of the endopiriform (EN) region of the claustrum complex, and understand their functional role. Here the authors have combined sophisticated intersectional viral tracing techniques, and ex vivo electrophysiology to map the neural circuitry of EN outputs to vCA1, and shown that optogenetic inhibition of the EN→vCA1 projection impairs both social and object recognition memory. Interestingly the authors find that the EN neurons target inhibitory interneurons providing a mechanism for feedforward inhibition of vCA1.

Strengths:

The strength of this study was the application of a multilevel analysis approach combining a number of state-of-the-art techniques to dissect the contribution of the EN→vCA1 to memory function.

Weaknesses:

Some authors would disagree that the vCA1 represents a 'node for recognition of familiarity' especially for object recognition although that is not to say that it might play some role in discrimination, as shown by the authors. I note however that the references provided in the Introduction, concerning the role of vCA1in memory refer to anxiety, social memory, temporal order memory, and not novel object recognition memory. Given the additional projections to the piriform cortex shown in the results, I wonder to what extent the observations may be explained by odour recognition effects. In addition, I wondered whether the impairments in discrimination following Chemo-genetic inhibition of the EN→vCA1 were due to the subject treating the novel and familiar stimuli as either both novel- which might be observed as an increase in exploration, or both stimuli as familiar, with a decrease in overall exploration.

Reviewer #2 (Public Review):

Summary:

Yamawaki et al., conducted a series of neuroanatomical tracing and whole-cell recording experiments to elucidate and characterise a relatively unknown pathway between the endopiriform (EN) and CA1 of the ventral hippocampus (vCA1) and to assess its functional role in social and object recognition using fibre photometry and dual vector chemogenetics. The main findings were that the EN sends robust projections to the vCA1 that colateralise to the prefrontal cortex, lateral entorhinal cortex, and piriform cortex, and these EN projection neurons terminate in the stratum lacunosum-moleculare (SLM) layer of distal vCA1, synapsing onto GABAergic neurons that span across the Pyramidal-Stratum Radiatum (SR) and SR-SML borders. It was also demonstrated that EN input disynaptically inhibits vCA1 pyramidal neurons. vCA1 projecting EN neurons receive afferent input from the piriform cortex, and from within EN. Finally, fibre photometry experiments revealed that vCA1 projecting EN neurons are most active when mice explore novel objects or conspecifics, and pathway-specific chemogenetic inhibition led to an impairment in the ability to discriminate between novel vs. familiar objects and conspecifics.

This is an interesting mechanistic study that provides valuable insights into the function and connectivity patterns of afferent input from the endopiriform to the CA1 subfield of the ventral hippocampus. The authors propose that the EN input to the vCA1 interneurons provides a feedforward inhibition mechanism by which novelty detection could be promoted. The experiments appear to be carefully conducted, and the methodological approaches used are sound. The conclusions of the paper are supported by the data presented on the whole.

However, some aspects of methodology and data interpretation will need to be clarified and further evidence provided to enhance the utility of the data to the rest of the field.

The authors used dual retrograde tracing and observed that the highest percentage (~30%) of vCA1 projecting EN cells also projected to the PFC. They then employed an intersectional approach to show the presence of collaterals in other cortical areas such as the entorhinal cortex and piriform cortex in addition to the PFC. However, they state that 'Projection to prefrontal cortex was sparse relative to other areas, as expected based on the retrograde labeling data' (referring to Figure 2K) and subsequently appear to dismiss the initial data set indicating strong axonal projections to the PFC.

Since this is a relatively unknown connection, it would be helpful if some evidence/discussion is provided for whether the EN projects to other subfields (CA3, DG) of the ventral hippocampus. This is important, as the retrograde tracer injections depicted in Figure 1B clearly show a spread of the tracer to vCA3 and potentially vDG and it is not possible to ascertain the regional specificity of the pathway.

The vCA1 projecting EN cells appear to originate from an extensive range along the AP axis. Is there a topographical organization of these neurons within the vCA1? A detailed mapping of this kind would be valuable.

Given this extensive range in the location of vCA1 EN originating cells, how were the targets (along the AP axis) in EP selected for the calcium imaging?

The vCA1 has extensive reciprocal connections with the piriform cortex as well, which is in close proximity to the EN. How certain are the authors that the chemogenetic targeting was specific to the EN-vCA1 connection?

Raw data for the sociability and discrimination indices should be provided so that the readers can gain further insight into the nature of the impairment.

Line 222: It is unclear how locomotor activity informs anxiety in the behavioral tests.

Figure 7 title; It is stated that activity of EN neurons 'predict' social/object discrimination performance. However, caution must be exercised with this interpretation as the correlational data are underpowered (n=5-8). Furthermore, the results show a significant correlation between calcium event ratios and the discrimination index in the social discrimination test but not the object discrimination test.

While both male and female mice were included in the anatomical tracing and recording experiments, only male mice were used for behavioral tests.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public Review):

Summary:

The anatomical connectivity of the claustrum and the role of its output projections has, thus far, not been studied in detail. The aim of this study was to map the outputs of the endopiriform (EN) region of the claustrum complex, and understand their functional role. Here the authors have combined sophisticated intersectional viral tracing techniques, and ex vivo electrophysiology to map the neural circuitry of EN outputs to vCA1, and shown that optogenetic inhibition of the EN→vCA1 projection impairs both social and object recognition memory. Interestingly the authors find that the EN neurons target inhibitory interneurons providing a mechanism for feedforward inhibition of vCA1.

Strengths:

The strength of this study was the application of a multilevel analysis approach combining a number of state-of-the-art techniques to dissect the contribution of the EN→vCA1 to memory function.

Weaknesses:

Some authors would disagree that the vCA1 represents a 'node for recognition of familiarity' especially for object recognition although that is not to say that it might play some role in discrimination, as shown by the authors. I note however that the references provided in the Introduction, concerning the role of vCA1 in memory refer to anxiety, social memory, temporal order memory, and not novel object recognition memory. Given the additional projections to the piriform cortex shown in the results, I wonder to what extent the observations may be explained by odour recognition effects.

We have added references demonstrating that the ventral hippocampus contributes to object recognition memory in rodents (Broadbent NJ et al., Learn Mem 2010; Titulaer J et al., Front Behav Neurosci 2021).

The odor recognition effect is an interesting perspective that we have also considered. However, in our object recognition test, the same odor (70% EtOH) was used for both objects, yet the mice were able to discriminate between the familiar and novel objects. This suggests that the likelihood of the odor cue contributing to their performance in object discrimination test is low.

In addition, I wondered whether the impairments in discrimination following Chemogenetic inhibition of the EN→vCA1 were due to the subject treating the novel and familiar stimuli as either both novel- which might be observed as an increase in exploration, or both stimuli as familiar, with a decrease in overall exploration.

We thank the reviewer for rising this interesting point. We analyzed the total exploration time (i.e., time in interaction zones in familiar and novel) during social discrimination test. The data is added to Fig. S9. Total exploration time was not affected by CNO treatment. This indicates inhibition of ENvCA1-proj. neurons reduced interaction time with the novel conspecific and increased interaction time with the familiar conspecific. The subject mice seem to give even weight on familiar and novel stimuli.

Reviewer #2 (Public Review):

Summary:

Yamawaki et al., conducted a series of neuroanatomical tracing and whole-cell recording experiments to elucidate and characterise a relatively unknown pathway between the endopiriform (EN) and CA1 of the ventral hippocampus (vCA1) and to assess its functional role in social and object recognition using fibre photometry and dual vector chemogenetics. The main findings were that the EN sends robust projections to the vCA1 that colateralise to the prefrontal cortex, lateral entorhinal cortex, and piriform cortex, and these EN projection neurons terminate in the stratum lacunosum-moleculare (SLM) layer of distal vCA1, synapsing onto GABAergic neurons that span across the Pyramidal-Stratum Radiatum (SR) and SR-SML borders. It was also demonstrated that EN input disynaptically inhibits vCA1 pyramidal neurons. vCA1 projecting EN neurons receive afferent input from the piriform cortex, and from within EN. Finally, fibre photometry experiments revealed that vCA1 projecting EN neurons are most active when mice explore novel objects or conspecifics, and pathway-specific chemogenetic inhibition led to an impairment in the ability to discriminate between novel vs. familiar objects and conspecifics.

This is an interesting mechanistic study that provides valuable insights into the function and connectivity patterns of afferent input from the endopiriform to the CA1 subfield of the ventral hippocampus. The authors propose that the EN input to the vCA1 interneurons provides a feedforward inhibition mechanism by which novelty detection could be promoted. The experiments appear to be carefully conducted, and the methodological approaches used are sound. The conclusions of the paper are supported by the data presented on the whole.

We thank the reviewer for their positive comments on our work.

The authors used dual retrograde tracing and observed that the highest percentage (~30%) of vCA1 projecting EN cells also projected to the PFC. They then employed an intersectional approach to show the presence of collaterals in other cortical areas such as the entorhinal cortex and piriform cortex in addition to the PFC. However, they state that 'Projection to prefrontal cortex was sparse relative to other areas, as expected based on the retrograde labeling data' (referring to Figure 2K) and subsequently appear to dismiss the initial data set indicating strong axonal projections to the PFC.

Our interpretation is that 70% of the ENCA1-proj. population does not send collaterals to the PFC, suggesting that the PFC is not a major target for this population (unlike vCA1 where 100% of its population projects). This hypothesis is supported by our axon branching study, which showed lower axon density in the PFC compared to vCA1 (and other regions). We revised the text to 'much sparser relative to that of vCA1' (line 101) to facilitate a direct comparison with the retrograde and anterograde labeling study.

Since this is a relatively unknown connection, it would be helpful if some evidence/discussion is provided for whether the EN projects to other subfields (CA3, DG) of the ventral hippocampus. This is important, as the retrograde tracer injections depicted in Figure 1B clearly show a spread of the tracer to vCA3 and potentially vDG and it is not possible to ascertain the regional specificity of the pathway.

We addressed the potential caveat associated with the retrograde tracer injection, as mentioned by the reviewer, by performing intersectional axon branching analysis. This analysis demonstrated that EN axons are primarily located in the SLM of the distal CA1 subfield (Figs. 2, 3, S2). However, we occasionally observed very weak labeling in the CA3 or dentate gyrus. We modified our text (lines 106-108) and figure (Fig. S2D) to account for this.

The vCA1 projecting EN cells appear to originate from an extensive range along the AP axis. Is there a topographical organization of these neurons within the vCA1? A detailed mapping of this kind would be valuable.

This is an interesting question for future research. Our data show a non-uniform distribution of this cell type, suggesting the potential for topographic organization.

Given this extensive range in the location of vCA1 EN originating cells, how were the targets (along the AP axis) in EP selected for the calcium imaging?

Using our injection coordinates, ENvCA1-proj. neurons were consistently labeled at high density just posterior to the bregma (Fig. 1J). Therefore, we targeted this region for our imaging.

The vCA1 has extensive reciprocal connections with the piriform cortex as well, which is in close proximity to the EN. How certain are the authors that the chemogenetic targeting was specific to the EN-vCA1 connection?

We performed histology on every animal used in the behavioral study to examine the specificity of hM4D expression, and only included those with specific labeling in the EN.

Raw data for the sociability and discrimination indices should be provided so that the readers can gain further insight into the nature of the impairment.

The raw data for total interaction time during the social discrimination test has been added (Fig. S9F).

Line 222: It is unclear how locomotor activity informs anxiety in the behavioral tests.

The degree of exploratory behavior in a novel context is generally considered to infer anxiety levels in rodents. We have added a review paper (Ref 44, Prut, 2003) that discusses this point.

Figure 7 title; It is stated that activity of EN neurons 'predict' social/object discrimination performance. However, caution must be exercised with this interpretation as the correlational data are underpowered (n=5-8). Furthermore, the results show a significant correlation between calcium event ratios and the discrimination index in the social discrimination test but not the object discrimination test.

We added the sample size for EN calcium imaging during the object recognition memory test (Fig. 7G). The updated data indicate a significant correlation between EN activity and the object recognition index (N = 9, Pearson R = 0.8, p = 0.01).

We have changed the title of Figure 7 to 'Activity of ENvCA1-proj. neurons correlates with social/object discrimination performance’.

While both male and female mice were included in the anatomical tracing and recording experiments, only male mice were used for behavioral tests.

The female behavior was highly inconsistent in the control condition of our social recognition memory paradigm; therefore, we decided to conduct the study with males. We will design a new behavioral paradigm for future studies to address this challenge.

Reviewer #1 (Recommendations For The Authors):

(1) It is not clear how the relative number of vCA1 projecting neurons in Figure 1H was acquired, not enough detail is presented in the methods section. To what extent could these data have been affected by differences in the size or anatomical position of the injection site in vCA1, which judging from the example fluorescent image in Figure 1B also appears to include CA3.

We used AMaSiNe (Song et al. 2020) to semi-automatically quantify fluorescently labeled presynaptic neurons. This open-source software identifies the number and location of these cells across different regions based on the Allen Mouse Brain Common Framework. To control for transfection variability (e.g., due to slight differences in injection volume or site), we normalized the presynaptic cell count in each region by the total number of cells in regions of interest. We performed for N = 5 brain and found consistent trend as seen in Fig. 1H (grey lines).

We have added the detailed method of quantification in the Materials and Methods section (line 393).

(2) For a number of the results, the full statistical values are not presented in the Results section or figure legend.

We have included the full statistical values in the figure legends of the revised manuscript.

(3) It is not clear how much virus was injected in the different experiments (tract racing, electrophysiology, behaviour, etc.). The methods state 50-100ul, but there is no further detail in the results or figure legends.

We have included the injected volumes of the virus in the revised manuscript.

(4) Figure 2 mentions the CLA complex (line 702) but this is not defined in the text. Although the introduction does refer to the claustrum complex, there is no acronym.

We have corrected the manuscript accordingly.

(5) Line 131- 'we recorded from 3-4 GABAergic neurons' - presumably this is in each animal?

We recorded 3 to 4 GABAergic neurons sequentially from the same slice to compare input strength. We have edited the text to clarify this (line 134).

Reviewer #2 (Recommendations For The Authors):

Figure 3C: It is not clear what the dashed lines labelled proximal and distal represent.

It is the proximal and distal vCA1 regions where GFP signals were measured for Fig. 3D. We have modified the figure legend to clarify this (line 736).

Figure 5D: what do the different colors represent? Different colors for one brain?

I assume that the reviewer meant to refer to Fig. 4D instead of Fig. 5D. In Fig. 4D, one color indicates starter cells in one brain. To clarify this, we have edited the figure legend (line 748).

Figure S6E: The images are low resolution and it is hard to decipher the exact locations of labeled neurons. Please provide more guidance (e/g/. labeling areas of interest).

We have added reference lines and labels in Figure S6E.

Some details are missing: what was the volume of AAV injected for each site/experiment; how was CNO made, and where was it purchased from?

We have added this information (lines 330-331; 431-434).

Version 2

Abstract

Abstract

The claustrum complex is viewed as fundamental for higher order cognition; however, the circuit organization and function of its neuroanatomical subregions are not well understood. We demonstrated that some of the key roles of the CLA complex can be attributed to the connectivity and function of a small group of neurons in its ventral subregion, the endopiriform (EN). We identified a subpopulation of EN neurons by their projection to the ventral CA1 (ENvCA1-proj. neurons), embedded in recurrent circuits with other EN neurons and the piriform cortex. Although the ENvCA1-proj. neuron activity was biased toward novelty across stimulus categories, their chemogenetic inhibition selectively disrupted the memory-guided but not innate responses of mice to novelty. Based on our functional connectivity analysis, we suggest that ENvCA1-proj. neurons serve as an essential node for recognition memory through recurrent circuits mediating sustained attention to novelty, and through feed forward inhibition of distal vCA1 neurons shifting memory-guided behavior from familiarity to novelty.

Introduction

The claustrum (CLA) complex, an evolutionarily conserved brain region found across mammalian species, reptiles, and birds, is hypothesized to serve as a node for establishing higher order cognitive functions by coordinating neuronal activity on a global scale (1-3). This view is based on its extensive connections with many cortical and subcortical areas (4, 5). Some of these functions include sensory perception and attention, which may affect memory processing, including working memory, associative memory, and recognition memory (6-8). Consistent with this view, abnormalities of the CLA complex is found in major cognitive disorders, including Alzheimer’s disease, schizophrenia, and attention-deficit/hyperactivity disorders (9-11). However, microcircuitry and functional segregation of the individual constituents of the CLA complex have remained unclear.

The cytoarchitecture and genetic expression broadly divide the CLA complex into dorsal and ventral parts, with the dorsal part typically referred to as “CLA” and ventral part having its own nomenclature: “endopiriform (EN)” in rodents (12, 13). Current evidence indicates that the CLA forms reciprocal connections with dorsal cortices containing sensory, motor and association areas generating diverse physiological effects depending on the cortical targets and their activity patterns (4, 5). In contrast, EN primarily provides inputs to the piriform cortex and limbic systems (14, 15), suggesting its distinct circuit organization and role in bridging olfactory information processed by the piriform cortex and memory processed by the limbic systems (16, 17).

One of distinct targets of EN, the ventral CA1 (vCA1), is known to coordinate in a range of behaviors related to exploration and recognition memory, and these functions are thought to be regulated by its afferent inputs in a domain-specific manner (18-26). We, therefore, hypothesized that EN is a key node for establishment of recognition memory. To address this, we set out to characterize circuit and function of EN neurons defined by their projection to vCA1 (EN vCA1-proj. neurons) using genetic tools and mouse models of social and object recognition memory.

We found EN vCA1-proj. neurons innervated multiple components of the limbic system except amygdala and prefrontal cortex and produced potent feedforward inhibitory control over vCA1 pyramidal neurons. During recognition memory test, activity of EN vCA1-proj. neurons were condensed around conspecifics or objects where mice spent most time on. However, disruption of EN vCA1-proj. activity only impaired memory-guided exploration of novel stimuli without affecting innate exploration induced by novelty. These findings demonstrate that EN subserves some of the key functions required for recognition memory governed by specific limbic system.

Results

Endopiriform represents a major afferent of the ventral CA1

To assess the significance of EN as a vCA1 afferents, we injected a retrograde tracer (60 nl) into the vCA1 of mice and compared the number of the retrogradely labeled neurons in the EN and other brain regions (Fig. 1A, B, S1A, Methods). We found labeled neurons in multiple areas including the septum, entorhinal cortex, and basolateral amygdala (Fig. 1C-F and Fig. S1B) (27, 28). In addition to these well-known vCA1 afferents, prominent labeling was observed in the area defined as EN in Allen Brain Atlas (Fig. 1G, Methods). The number of presynaptic neurons in EN was ∼10 fold lower than entorhinal cortex (data not shown). However, the normalized count of presynaptic neurons in EN was significantly greater than basolateral amygdala, lateral septum, and medial septum (Fig. 1H).

Endopiriform projects to the ventral CA1

(A) Schematic diagram of a retrograde tracer injection performed. CAV-2-Cre or CTB647 was used in combination with Ai14 or wild-type mice, respectively. (B) An example fluorescent image of injection site. (C) Distribution of retrogradely labeled vCA1-projecting neurons in entire brain as a result of injection in panel B. (D-G) Fluorescent images showing retrogradely labeled vCA1-projecitng neurons in septum (SP), entorhinal cortex (ENT), basolateral amygdala (BLA), and endopiriform (EN). Str: Striatum, CLA: Claustrum. (H) Box plots showing relative number of vCA1-projecting neurons in different brain areas. Cell number in each brain area were normalized by summed cell number of EN, BLA, SP (lateral), and SP (medial). Cell numbers in each brain areas of individual brains are presented in Table 1. Median value for EN, BLA, SP (lateral), and SP (medial) is 0.45, 0.24, 0.08, and 0.21, respectively. Statistics: EN vs BLA, SP (lateral), or SP (medial), p < 0.001 for all; BLA vs SP (lateral) or SP (medial): p = 0.002 and 0.694, respectively; SP (lateral) vs SP (medial) p = 0.023. One-way ANOVA, repeated comparison with Tukey-Kramer test. N = 5. (I) An example fluorescent image of horizontal section showing retrogradely labeled vCA1-projecting neurons in EN. A: anterior, P: posterior. (J) Top: Example map showing distribution of vCA1-projecting neurons in entire EN in one mouse. Bottom left: A plot showing a quantified distribution at bin size of 1 mm. Each bin was normalized to total number of labeled neurons in EN. Data from each mouse is shown in grey. Median is indicated in black. Bottom right: Box plots comparing a number of vCA1-projecting neurons in anterior and posterior EN. Data were normalized to a total number of labeled neurons in EN. Median value for anterior vs posterior was 0.39 vs 0.54. Statistics: p = 0.008, Rank-sum.

The vCA1-projecting neurons in the EN were spread along antero-posterior axis with two peak densities, one in anterior and another in posterior to the bregma (Fig. 1I, J). Number of labeled neurons was greater in posterior EN (Fig. 1J). In contrast, the same analysis along dorso-ventral axis showed a single dominant peak at the depth consistent with the location of EN (Fig. S1C). To further confirm the location of these vCA1-projecting neurons, we injected a retrograde tracer (60 nl for each) in one color into vCA1 and a retrograde tracer in another color into a cortical area known to receive projections from EN or “CLA” representing a dorsal part of CLA complex (Fig. 2A). We targeted the prefrontal cortex to label EN, and the anterior cingulate cortex, the motor cortex, and the dorsomedial entorhinal cortex to label CLA (Fig. 2C-F) (29-32).

Axon branching of vCA1-projecting EN neurons

(A) Schematic diagram of injection performed and two hypothesized outcomes. Yellow color indicates double-labeled neurons. PFC: Prefrontal cortex, ACC: Anterior cingulate cortex, MC: Motor cortex, ENT(dm): Dorsomedial entorhinal cortex, CLA: claustrum (B-F) Example fluorescent images of injection site for each region indicated. (G-J) Example fluorescent images of retrogradely labeled neurons in CLA complex. vCA1-projecting neurons were colored in red and cortex-projecting neurons were colored in green. (K) Box plots showing percentage of double-labeled vCA1-projecting neurons for each cortical injection. Medial value for PFC, ACC, MC, and ENT(dm) was 28.6, 4.6, 0, and 1.2, respectively. Statistics: PFC vs ACC, MC or ENT(dm), p = 0.405, 0.015 and 0.047, respectively. ACC vs MC or ENT (dm), p = 0.422 and 0.64, respectively. MC vs ENT (dm), p = 0.993. One-way ANOVA with post-hoc Kruskal-Wallis test. Slice and animal number: 16 and 4 for PFC, 20 and 5 for ACC, 10 and 3 for MC, and 18 and 4 for ENT(dm). (L) Schematic diagram of projection pattern found in the experiment shown in A. (M) Schematic diagram of the intersectional approach performed to label vCA1-projecting EN neurons. (N-O) Example fluorescent images showing vCA1-projectig axons in different brain areas. All images were from the same mouse. (P) Magnified images showing vCA1-projecting axons in cortical layers. Magnified areas are indicated in O and fig. S2A-C as dashed squares. PIR: Piriform cortex

We found there is a major spatial overlap in labeling of vCA1-projecting neurons with prefrontal cortex-projecting neurons, but not with other cortical projection neurons (Fig. 2G-J). Moreover, around 30% of vCA1-projecting neurons were found to be double-labelled when a cortical injection was made into the prefrontal cortex, suggesting some vCA1-projecting neurons send axons to this cortical area (Fig. 2K, L).

To further determine the cortical and subcortical innervation pattern of vCA1-projecting EN neurons, we used an intersectional approach to label their somata and axons with eGFP. CAV2-Cre (60 nl) was injected into the vCA1 and flox-eGFP (100 nl) was injected into the EN (Fig. 2M). Apart from vCA1, axons from vCA1-projecting neurons were found in various cortical areas including prefrontal cortex, lateral entorhinal cortex, and piriform cortex (Fig. 2N-P and Fig. S2A-C). Projection to prefrontal cortex was much sparser relative to that of vCA1, as expected based on the retrograde labeling data (Fig. 2K). The connectivity of vCA1-projecting EN neurons to the amygdala, which represents the major component of limbic systems, was also sparse relative to vCA1 (Fig. S2C), indicating vCA1 is the main target of these EN neurons in this system. Within the hippocampus, very sparse EN axons were observed in vCA3 and dentate gyrus (Fig. S2D).

Taken together, these findings indicate EN represents a major afferent of vCA1. Moreover, EN neurons projecting their axons to vCA1 also send strong collaterals to lateral entorhinal cortex and piriform cortex, but relatively weak to the prefrontal cortex or other limbic structures. Since our subsequent study focused on EN neurons defined by their projection to vCA1, we refer to them as ENvCA1-proj. neurons.

Projection pattern and synaptic connectivity of EN axons in vCA1

We next investigated projection pattern of EN axons in hippocampus by injecting an anterograde tracer (AAV-GFP, 60 nl) into the EN (Fig. 3A, B). Analysis of different parts of hippocampal section indicated EN axons were mostly confined in SLM layer of distal part of vCA1 (Fig. 3C, D). Projection to intermediate or dorsal CA1 was limited or undetectable (Fig. S3A-C).

EN projection pattern in vCA1

(A) Schematic diagram of injection performed. (B) An example fluorescent image of horizontal brain section containing injection site in EN. (C) An example fluorescent image of ventral hippocampal slice with EN axons resulted from the injection in panel B. Dashed lines indicate the distal and proximal vCA1 regions where GFP signals were measured for the plot in panel D. (D) Normalized GFP signals across the layer. All signals were normalized by maximum GFP signal in the distal subregion. Median of area under the curve: distal; 11.12, proximal; 5.99. distal vs proximal, p = 0.016. Wilcoxon signed-rank test. n = 8 slices (N = 4 mice). SR: stratum radiatum, Py: pyramidal layer.

Since SLM mainly consists of GABAergic neurons (33), we hypothesized that EN axons form synapses with this cell type. To test this, we first applied monosynaptic rabies tracing technique to Vgat-Cre mice (Fig. 4A, Methods). The distribution of starter cells (identified by GFP and mCherry co-labeling) was confirmed to be within the ventral-intermediate region of hippocampus using AMaSiNe (Fig. 4B, S4A) (34). In these mice, the presynaptic neurons were consistently observed in EN in addition to other expected areas (e.g., septum and entorhinal cortex) (Fig. 4C and D, S4B-F) (35, 36). The density of the presynaptic neurons in EN was greater at the posterior region to the bregma (Fig. 4D), consistent with our earlier observation (Fig. 1J).

ENvCA1-proj. neurons target vCA1 interneurons

(A) Schematic diagram of injection performed for monosynaptic rabies tracing. Flox-TVA-GFP (60 nl) was injected into the vCA1. 4 weeks later, Rabies-mCherry (60 nl) was injected. (B) Location of starter cells in the hippocampus. Data from 5 brains (each brain represented by different color) were overlaid. (C) Example images showing presynaptic cells in EN (white arrow). (D) Top: Location of presynaptic EN neurons. Data from 5 brains were overlaid (different brains were coded by different colors). Bottom: Plots indicate distribution of presynaptic cells in EN along AP axis. Cell counts in a given bin (1 mm) were normalized by total cell number in EN. Thin gray lines indicate individual data. A thick black line indicates median. Inset showing plot comparing cell number in anterior and posterior location as in Fig. 1J. Median: anterior vs posterior, 0.25 vs 0.72, p = 0.031; Wilcoxon signed-rank test. (E) Top: Schematic diagram of injection performed. Bottom: An example fluorescent image of injection site in EN. (F) Schematic diagram of recordings performed. Recordings were performed from identified GABAergic neurons sequentially in a random order. (G) An example confocal image of an acute vCA1 slice used for recordings. Recorded GABAergic neurons were filled with biocytin and post-labeled with Alexa488. White dashed lines indicate border of Py-SR and SR-SLM. (H) Median traces of photo-evoked EPSCs recorded from neurons at different laminar position. (I) Connection probability (CP) of EN axons and recorded neurons at different laminar positions. Numbers in bars indicate number of response positive neurons /total recorded neurons. (J) Left: Peak amplitudes of evoked EPSCs (normalized to total input from recorded slice). Right: the same as left but for mean amplitude. Gray circles indicate each neuron. Black lines indicate median. Median (peak amp): Py-SR vs SR, SR-SLM or SLM: 1.1 vs 1.0 vs 1.1 vs 0.0. [Py-SR vs SR, SR/SLM vs SLM], p = 0.999, 0.988, 0.076; [SR vs SR-SLM or SLM], p = 0.972, 0.111; [SR-SLM vs SLM], p = 0.039; one-way ANOVA, repeated comparison with Tukey-Kramer test. Median (mean amp): Py-SR vs SR, SR-SLM, or SLM: 1.0 vs 0.8 vs 1.4 vs 0.0. [Py-SR vs SR, SR-SLM or SLM], p = 0.991, 0.999, and 0.141; [SR vs SR-SLM or SLM], p = 0.978, and 0.251; [SR-SLM vs SLM], p = 0.114; one-way ANOVA, repeated comparison with Tukey-Kramer test. N = 5 mice.

To determine how EN inputs are integrated into the local circuit in vCA1, we systematically assessed the connections between EN axons and GABAergic neurons in different layers of distal vCA1. We expressed channelrhodopsin-2 (ChR2) into the EN axons by injecting AAV-ChR2-mCherry (100 nl) into EN and performed whole-cell recording from GABAergic neurons in acute vCA1 slices in the presence of TTX (1 µM) and 4-AP (100 µM) to isolate monosynaptic connection (37). To ensure the recording from GABAergic neurons, we used Vgat::Ai14 mice in which all GABAergic neurons are labeled with tdTomato (Fig. 4E-G). We recorded from 3 to 4 GABAergic neurons located in different laminar position in distal vCA1 (Py-SR border, SR, SR-SLM border, or SLM) in the same slice to construct a “laminar profile” of EN◊vCA1 inputs (Fig. 4F and G). Whole-field LED stimulation evoked excitatory postsynaptic current (EPSC) in most GABAergic neurons recorded. However, contrary to the expectation, connection probability and strength were lowest in neurons in SLM compared to neurons in other layers (Fig. 4H-J). Taken together, these results indicate EN axons preferentially innervate the GABAergic neurons spanning across the Py-SR border to SR-SLM border in distal vCA1.

EN axons produce feedforward inhibition onto vCA1 pyramidal neurons

SLM also contains the apical tuft dendrite of pyramidal neurons, which could be a target of EN axons in addition to GABAergic neurons. To test this, we expressed ChR2 in EN axons but made whole-cell recording from pyramidal neurons in distal vCA1 in the presence of TTX and 4-AP. Since CA1 pyramidal neurons can be differentiated into superficial and deep layer neurons with distinct anatomy and function (38), we recorded from neurons in each layer as well as nearby GABAergic neurons at the Py-SR and SR-SLM borders for the comparison (Fig. 5A). We found monosynaptic input to superficial and deep layer pyramidal neurons was similar, and monosynaptic input to pyramidal neurons (superficial and deep pooled) and GABAergic neurons was also comparable (Fig. S5).

EN disynaptically inhibits vCA1 pyramidal neurons

(A) A confocal image of recorded pyramidal neurons in vCA1. Recorded pyramidal neurons were filled with biocytin and post-labelled with Alexa488. Ori.: oriens, Super.: superficial layer pyramidal neuron. Deep: deep layer pyramidal neuron. (B) Example traces of photo-evoked EPSCs and IPSCs in superficial (black trace) and deep (gray trace) pyramidal neurons. Red trace: after TTX and 4AP. Blue trace: after gabazine. (C) Scatter plots of peak amplitudes of EPSCs and IPSCs for each neuron. Dashed line indicates a unitary line. Superficial neurons: empty Black circles. Deep neurons: Gray circles. N = 4 mice. (D) A summary diagram of EN◊vCA1 circuit. In vCA1, EN axons innervate GABAergic neurons (IN), which in turn inhibits pyramidal neurons (both superficial and deep).

Since EN axons innervated multiple GABAergic neurons across different layers, we hypothesized they may exert stronger inhibition onto pyramidal neurons through feed-forward inhibition. To test this, we recorded EPSC and inhibitory postsynaptic current (IPSC) from superficial and deep pyramidal neurons using cesium-based internal solution in the brain slices expressing ChR2 in the EN axons (Methods). At the command potential of -70 mV, photo-stimulation of EN axons evoked EPSCs in majority of pyramidal neurons in both layers (7 out of 10 neurons in superficial and 5 out of 10 neurons in deep) (Fig. 5B). At the command potential of +12 mV, the same stimulation evoked outward postsynaptic current in the same neurons, including those without EPSCs (10 out of 10 for superficial and deep) (Fig. 5B). This outward current was abolished by application of TTX and 4-AP or gabazine (10 µM), indicating it is disynaptically driven and mediated by inhibitory GABAA receptor (Fig. 5B). When relative strength of EPSCs and IPSCs were compared for each neuron, the strength of IPSCs overwhelmed the EPSCs in all cases (Fig. 5C). This indicates EN axons disynaptically inhibit pyramidal neurons in vCA1 (Fig. 5D).

ENvCA1-proj. neurons receive inputs from the piriform cortex

We next examined afferents that may drive the EN◊vCA1 circuit. To address this, retrograde CAV2-Cre was injected into vCA1 to express Cre into ENvCA1-proj. neurons, then applied monosynaptic rabies tracing technique to specifically label its presynaptic neurons in wild type mice (Fig. 6A). Starter cells and presynaptic neurons were spatially mapped and quantified using AMaSiNe (Fig. 6B and C, S6). Since presynaptic neurons were sparse in contralateral hemisphere, data from ipsilateral hemisphere was analyzed. Moreover, we applied minimum labeling threshold to determine the brain region for further analysis (Methods). We found the piriform cortex contained highest number of presynaptic neurons followed by EN (Fig. 6D-F). Within piriform cortex, layer 2 neurons were the major source of afferent (Fig. 6G). These data indicate ENvCA1-proj. neurons receive major input from piriform cortex and form a recurrent connection with neurons in EN (Fig. 6H).

ENvCA1-proj. neurons receive inputs from the piriform cortex and within EN

(A) Schematic diagram of performed. After CAV2-Cre (60 nl) injection into the vCA1, flox-TVA-mCherry (100 nl) was injected into the EN. 4 weeks later, Rabies-GFP (100 nl) was injected into the EN. (B) Location of starter cells in EN. Data from 6 brains were overlaid. (C) An example map of presynaptic cells in one brain. Top: Horizontal view. Injections were made in left hemisphere. Bottom: Front view. (D) Number of presynaptic cells in different brain areas (normalized to total presynaptic cell number). Median for AIA, AON, PIR, ENT(lateral), CLA, EN, Pallidum, and SI (in fraction): 0.7, 0.3, 3.8, 0.3, 0.5, 1.7, 0.3, and 0.2. [AON, ENT(lateral), CLA, Pallidum, SI] vs PIR: p = 0.016, 0.008, 0.048, 0.006, and 0.003. EN vs SI, p = 0.043. one-way ANOVA with post-hoc Kruskal-Wallis test. Full list of statistical tests and p values are in Table 2. (E) An example fluorescent image of coronal slice containing EN and PIR and labeled presynaptic cells. (F) Example maps of presynaptic cells in PIR (top) and EN (bottom). (G) Number of presynaptic cells in PIR in different layers (normalized to the total number of labeled neurons in PIR). Median: L1 vs L2 vs L3, 0.11 vs 0.61 vs 0.27. L1 vs L2, p < 0.001; L1 vs L3, p = 0.005; L2 vs L3, p < 0.001; one-way ANOVA, repeated comparison with Tukey-Kramer test. (H) Summary diagram showing input-output circuit of ENvCA1-proj. neurons (red circle).

The activity of ENvCA1-proj. neurons is correlated with the time mice spend in space

The downstream target of EN, vCA1, is implicated in social, odor and object recognition memory (20-22). To determine whether and how the activity of EN neurons is related to this function, we expressed GCaMP8s in ENvCA1-proj. neurons and monitored their activity using fiber photometry while video recording mice performing social recognition memory test consisting of pretest, sociability test, and discrimination test (Fig. 7A, B, Methods). The nose position of the subject mouse was tracked during tests using DeepLabCut (Fig. 7C) (39, 40), and cumulative time and calcium event were mapped onto the arena space (Fig 7C, S7). We found these values were correlated in space across all sessions (i.e., the more time the mice spent in a given space, the more the ENvCA1-proj. activity), including in the open field with no object or in the test arena with conspecifics or objects (Fig. S7, S8A-B).

Activity of ENvCA1-proj. neurons correlates social/object discrimination performance

(A) Right and Middle: Schematic diagrams of injection and optic cannula implantation performed. CAV2-Cre (60 nl) was injected into the vCA1, and dlox-jGCaMP8s (100 nl) was injected into the EN. Left: An example image showing jGCaMP8s expressing ENvCA1-proj. neurons and a tract of optic cannula. (B) The diagram of social recognition memory test. A test consisted of pretest (5 min), sociability test (10 min), and discrimination test (5 min). (C) Left: An example image of a mouse with nose point marker tracked by DLC. Right: An example map of nose points and calcium events in the arena during sociability test. (D) Nose distance vector (distance between subject’s nose point and a center point of novel/familiar mouse chambers) and simultaneously acquired calcium signals during discrimination test. Gray dashed lines indicate the border of interaction zones. Open red circles indicate the calcium event detected (threshold: Z > 2.58). A part of calcium signal trace was expanded at the bottom for clarity. (E) Left: Correlation between calcium event ratios (calcium event number in the mouse interaction zone/ calcium event number in the object interaction zone) and sociability index. p = 0.29, N = 8 mice. Right: same as in left panel, but for calcium event ratios (calcium event number in the unfamiliar mouse interaction zone/ calcium event number in the familiar mouse interaction zone) and discrimination index. p = 0.003, N = 8. (F) The diagram of novel object recognition memory test. A test consisted of habituation (5 min), familiarization (10 min), and discrimination test (5 min). (G) Same as in (E), but for calcium event number ratios (calcium event number in the unfamiliar object interaction zone/calcium event number in the familiar object interaction zone and discrimination index. p = 0.01. N = 9 mice.

Cumulative calcium events of ENvCA1-proj. neurons became high towards the space around objects or conspecifics when they are present and was correlated with interaction times except for sociability session (Fig. S8C-D). In the sociability session, the cumulative calcium events around a novel conspecific and object were similar despite mice spending more time on conspecifics (Fig. 7E left, S8C). In contrast, in the social discrimination test, both the cumulative time and cumulative ENvCA1-proj. activity were higher for an unfamiliar than familiar conspecific (Fig. 7D-E, S8C). As a result, ENvCA1-proj. activity and behavior had a greater degree of correlation in social discrimination than sociability.

We also tested if a similar correlation occurs in a non-social context by recording ENvCA1-proj. activity and behavior in the object discrimination test (Fig. 7F). The cumulative time and calcium events were significantly correlated across different sessions i.e., familiarization, and object discrimination (Fig. 7G, S8E), largely consistent with the findings in the social discrimination test.

These data indicate that the activity of ENvCA1-proj. neurons is generally correlated with the time the mouse spends in a particular location at the basal level (i.e. open field) or task conditions. However, the degree of correlation appears to vary depending on the session, such that it was highest in the pretest/familiarization and social/object discrimination test, but lowest in the sociability test.

Inhibition of ENvCA1-proj. neurons impairs social/object recognition memory

Although ENvCA1-proj. activity predicted the behavior in pretest/familiarization and discrimination test better than sociability test, their causality is unclear. To address this, we inhibited the activity of ENvCA1-proj. neurons using inhibitory DREADD (hM4Di) during social and object discrimination tests (Fig. 8A-C and h, S9A) (41). To specifically express hM4Di in ENvCA1-proj. neurons, CAV-2-Cre (60 nl) and AAV-flox-hM4Di (100 nl for each coordinate) were bilaterally injected into vCA1 and EN, respectively (Fig. 8A-B). hM4Di was replaced with tdTomato for control group. We made a protocol that allows within-subject comparison, such that a mouse goes through a test with chlozapine-N-oxide (CNO, 1 mg/ kg) treatment then the same test with saline treatment the next day, or vice versa (Fig. 8C and H). In social discrimination test, CNO treatment did not affect the pretest or sociability in hM4Di or control group, but specifically impaired social discrimination in hM4Di group (Fig. 8D-G, S9B-C). A similar effect was also observed in novel object discrimination (Fig. 8I, S9D).

Chemogenetic inhibition of ENvCA1-proj. neurons specifically impairs recognition memory

(A) Schematic diagram of injection performed. (B) Left: An example fluorescent image showing expression of hM4Di in ENCA1-proj. neurons. Right: Expansion of areas marked by white rectangle in image on the left. (C) The diagram of social recognition memory test. A session consisted of habituation (5 min), Pretest (5 min), Sociability test (10 min), and discrimination test (5 min). The group treated with a saline in session 1 was treated with CNO (1mg/ kg) in session 2 the following day. (D) Example cumulative times heat map in arena during sociability tests for the same mouse under two different treatments. (E) Plots comparing sociability index for hM4Di and tdTomato mice under two different treatments. Gray empty circles indicate each mouse. Gray thin lines show paired data from one mouse. Thick black bars indicate median. Median sociability index: hM4Di mice (Saline vs CNO), 0.75 vs 0.76, p = 0.391; signed-rank test; tdTomato mice (Saline vs CNO), 0.77 vs 0.75, p = 0.383; signed-rank test. (F) Example cumulative time heat maps in arena during discrimination tests for the same mouse under two different treatments. (G) Comparison of discrimination index for hM4Di and tdTomato mice under two different treatments. Gray empty circles indicate each mouse. Gray thin lines show paired data from one mouse. Thick black bars indicate median. Median discrimination index: hM4Di mice (Saline vs CNO), 0.66 vs 0.57, p = 0.042; signed-rank test; tdTomato mice (Saline vs CNO), 0.65 vs 0.63, p = 0.742; signed-rank test. (H) The diagram of novel object recognition memory test. A session consisted of habituation (5 min), familiarization (10 min), and discrimination test (5 min). CNO treatment was the same as in C. (I) Same as in G, but for novel object discrimination index. Median discrimination index: hM4Di mice (Saline vs CNO), 0.64 vs 0.5, p = 0.003; signed-rank test.

Apart from recognition memory, vCA1 is implicated in anxiety and associative fear memory (42, 43), thus, we tested contribution of ENvCA1-proj. neurons to these functions. The anxiety level as determined by locomotor activity (44) and sociability was not affected by CNO treatment in hM4Di group (Fig. 8E, S9E). Similarly, fear memory as assessed by the freezing response during training (with trace fear conditioning) or recall to context or tone was not affected by CNO treatment in hM4Di group (Fig. S10).

Taken together, these data indicate that ENvCA1-proj. activity is not causally related to innate exploration behavior induced by novelty, or anxiety or fear memory. However, their activity is necessary for mice to discriminate between familiar and unfamiliar conspecific or object, suggesting their major role in general recognition memory.

Discussion

We characterized the circuit and function of EN neurons targeting specific subregions and layers of vCA1. In contrast to other parts of CLA complex, ENvCA1-proj. neurons were reciprocally connected with the piriform cortex representing a major downstream target of the olfactory bulb (45). This circuit motif may predict the predominant role of ENvCA1-proj. neurons in social recognition memory, given the ability of social odor to engage multiple olfactory pathways innervating the piriform cortex (46). However, we found that these neurons contributed to broader domains of recognition memory. This is supported by the observations that the inhibition of ENvCA1-proj. activity impaired both social and non-social discrimination.

In all phases of recognition memory test, EN vCA1-proj. activity was highly correlated with the time mice spent in given locations and this correlation was biased towards novel stimuli when mice were engaged in their exploration irrespective of the stimuli being social or non-social (Fig. 9A). Taken together, these data suggest that the function of ENvCA1-proj. neurons was primarily related to the detection of and attention to novel stimuli. The recurrent connections within EN, which is another prominent feature of ENvCA1-proj. neurons’ network, potentially support these processes to help establish the recognition memory (47).

The role of ENvCA1-proj. neurons in recognition memory.

(A) Summary of findings from in vivo experiments. (B) The model of ENvCA1-proj. circuit function in recognition memory. During memory-guided exploration, recurrent circuits in ENvCA1-proj. neurons maintain the attentional response to novel stimuli while regulating the degree of response to familiar stimuli in vCA1 via feedforward inhibition.

Attentional processes mediated by ENvCA1-proj. neurons can be considered of particular relevance for memory-guided behavior, given that inhibition of ENvCA1-proj. activity selectively impaired conspecific/object discrimination but not innate exploratory behavior provoked by novelty as seen in pretest, sociability, and familiarization (Fig. 9A). This functional specialization, likely associated with its unique circuit connectivity to the limbic system, could differentiate EN from CLA.

Studies on EN are generally scarce, and the available evidence of EN projections to vCA1 suggests that these projections are sparse (14). Potential reasons include the difficulty in clearly delineating the EN from CLA complex and non-uniform distribution of ENvCA1-proj. neurons along antero-posterior axis. By leveraging contemporary circuit analysis tools, we demonstrated that the EN◊vCA1 circuit is non-trivial; ENvCA1-proj. neurons were more numerous than commonly studied vCA1 afferents, including those from the septum and amygdala. Moreover, these axons exhibited a potent disynaptic inhibition of distal vCA1 pyramidal neurons, indicating their major functional implications for vCA1 activity.

Our data indicated that although EN axons terminated at the SLM of distal vCA1, they exhibited stronger connections to GABAergic neurons spanning from the Py-SR border to the SR-SLM border than to SLM neurons. The detection of EN inputs from GABAergic neurons at the Py-SR border was surprising, considering the distance between soma position and the EN axons. However, there are several GABAergic cell types below SLM that extend their primary dendrites into SLM (48, 49), likely explaining the observed connectivity pattern. Among these are chandelier cells found at the Py-SR border, which preferentially inhibits the axon initial segments of pyramidal neurons, and are thus well suited to exert the powerful feedforward inhibition of vCA1 pyramidal neurons observed in our recordings (50).

There are several potential circuit effects by which EN inputs could affect vCA1 function. One is the induction of non-linear events in vCA1 pyramidal neurons by acting in concert with other afferents (e.g., lateral entorhinal cortex targeting SLM) (36) to generate a specific output pattern in pyramidal assemblies representing task-relevant stimuli. Alternatively, EN inputs may improve signal to noise ratio of the information conveyed through the lateral entorhinal cortex, CA2 (in the case of social context), and possibly other afferents (e.g., CA3) (51). Another scenario is tuning of vCA1 pyramidal neurons’ response to familiar stimuli through dynamic combination of a weak monosynaptic excitatory input and strong disynaptic inhibitory input. Such mechanism suggests that at a milliseconds scale novelty (EN) and familiarity (vCA1) recognition might occur though alternating rather than simultaneous activity of brain microcircuits. Taken together, we propose a model for the role of ENvCA1-proj. neurons in recognition memory by balancing memory-guided attentional responses to familiarity and novelty through combination of feedforward inhibition of vCA1 (a node for recognition of familiarity) and the recurrent circuits that contribute to sustain attention to novelty (Fig. 9B).

The model of promoting novelty detection by suppressing familiarity responses is consistent with previous observations showing that vCA1 pyramidal neurons predominantly respond to familiar stimuli whereas novelty response can only occur through activation of vCA1 interneurons (52, 53). Nevertheless, alternating activation of EN and vCA1 could prove essential for the behavioral relevance of EN activity associated with recognition memory, without affecting innate exploratory behavior to novelty. Accordingly, the latter behavior is primarily associated with the prefrontal cortex and amygdala, areas that are only weakly targeted by ENvCA1-proj. neurons (Fig. 2P, S2C) (54-56).

In addition to advancing our understanding of the basic organization and function of brain circuits underlying higher cognitive processes, our findings suggest dysfunction of ENvCA1-proj. neurons could be a key contributing factor to the deficits in CLA complex function and recognition memory found in neuropsychiatric disorders (9, 10). Discrete EN populations can thus emerge as pathophysiological substrate, but also as important treatment targets of key symptoms of these disorders.

Materials and methods

Mice

All experiments were performed in accordance with standard ethical guidelines and were approved by the Danish national animal experiment committee (License number: 2021-15-0201-00801). Unless otherwise noted, all mice used were C57BL6/J mice or transgenic mice with C57BL6/J backgrounds. They were 2-4 months old at the time of the experiment. Transgenic mice used were B6.Cg-Gt (ROSA)26Sortm14(CAG-tdTomato)Hze/J mice (Ai14, JAX007914) and B6J.129S6(FVB)-Slc32a1tm2(cre)Lowl/MwarJ mice (Vgat-Cre, JAX028862). Similar number of male and female mice were used for all experiments except for behavior tests. For behavior tests, male mice were used.

Viruses and retrograde tracers

The information on viruses used in our experiments was the following. CAV-2-Cre (PVM); AAV1-mDlx-HBB-chl-dlox-TVA_2A_oG(rev)-dlox (v271-1, VVF); AAV1-hSyn1-dlox-TVA_2A_mCherry_2A_oG(rev)_dlox (v306-1, VVF); Rabies-GFP (NTNU viral core facility); Rabies-mCherry (NTNU viral core facility); AAV5-CAG-GFP (37525-AAV5, Addgene); AAV5-CAG.hChR2(H134R)-mCherry (10054-AAV5, Addgene); AAV8-hSyn-DIO-hM4Di (Gi) – mCherry (44362-AAV8, Addgene); AAV5-hEF1a-dlox-EGFP(rev)-dlox (v217-5, VVF); and retrograde AAVrg-CAG-GFP (37825-AAVrg, Addgene), AAV1-hSyn1-dlox-jGCaMP8s(rev) (v627-1, VVF). We also used retrograde traces of red Retrobeads (Lumafluor) and Cholera toxin subunit B conjugated with Alexa 647 (CTB647, Thermo Fisher).

Stereotaxic injections

Stereotaxic injections were performed using a stereotaxic frame (Model 940, Kopf). Mice were anesthetized with isoflurane and were subcutaneously injected with buprenorphine (0.1 mg/kg) and metacam (1.5 mg/Kg) for post-operative pain relief. After incising the scalp over the cranium, a small hole was bored with a microdrill bit, and a beveled glass pipette (Wiretrol II, 5-000-2010, Drummond Scientific Company) back-filled with mineral oil and front-filled with the material to be injected was slowly inserted into a target coordinate. After injecting a small volume (50 - 100 nL) with a custom-made displacement injector (based on MO-10, Narishige), the pipette was left in place for 3 to 5 mins before slow retraction. The incision was closed with a nylon suture. The stereotaxic coordinates used were (relative to bregma, in mm): anteroposterior (AP) -3.0; mediolateral (ML) 3.12; dorsoventral (DV) 3.7 for ventral CA1; AP +2.0; ML 0.2; DV 1.5 and 2.5 for prefrontal cortex; AP 0.0; ML 0.2; DV 1.0 and 1.5 for anterior cingulate cortex; AP 0.0; ML 1.5; DV 0.3 and 0.7 for motor cortex; and AP -5.0; ML 3.0; DV 1.5 for dorsomedial entorhinal cortex. For endopiriform injection, we injected two sites from the following coordinates: AP -0.27, ML 3.2, DV 4.2; AP 0.0, ML 3.0, DV 4.2; or AP +1.0, ML 2.7, DV 4.0. Mice were thermally supported with a feedback-controlled heating pad maintained at ∼37 °C (ThermoStar Homeothermic system, RWD). Mice were used for experiments 3-5 weeks post injection.

Implantation of the optic probe

After stereotaxic injection of CAV-2-Cre into the vCA1 and AAV1-hSyn1-dlox-jGCaMP8s into the EN (AP -0.27, ML 3.2, DV 4.2), a fiber optic cannula (400 μm core diameter, 4.5 mm length, NA 0.39, R-FOC-BF-400C-39NA, RWD) was inserted towards EN. Once in the correct depth, exposed brain surface around the implant was covered by Kwik-Cast (WPI) and was then fixed on skull with dental adhesive resin cement (Super-Bond, SUN MEDICAL).

Ex vivo electrophysiology

Mice were deeply anesthetized with isoflurane and decapitated. Horizontal sections (300 µm) containing ventral hippocampus were prepared by vibratome (VT1200S, Leica) in ice-cold choline solutions containing (in mM): 25 NaHCO3, 1.25 NaH2PO4-H2O, 2.5 KCl, 0.5 CaCl2, 7 MgCl2, 25 D-glucose, 110 Choline chloride, 11.6 Ascorbic acid, and 3.1 C3H3NaO3. Slices were subsequently incubated in artificial cerebrospinal fluid (ACSF) containing (in mM): 125 NaCl, 25 NaHCO3, 1.25 NaH2PO4-H2O, 2.5 KCl, 11 D-glucose, 2 CaCl2, and 1 MgCl at 34 °C for 30 mins then at room temperature (∼20 °C) for at least 1 hour before the recording.

Whole-cell recording was performed with an upright microscope (BX51WI, Olympus) equipped with a motor-controlled stage and focus (MP285A and MPC-200, Sutter instrument), differential interference contrast, coolLED (model pE300), and a monochrome camera (Moment, Teledyne). Neurons were visualized with a 60x lens (1.00 NA, LUMPlanFL N, Olympus) with the software Micro-Manager-2.0 gamma (US National Institutes of Health). Pipettes (4-5 MΩ) were pulled from thick-walled borosilicate capillary glass with a puller (model P-1000, Sutter instrument). For a voltage-clamp recording, the internal solution contained (in mM): 135 CsMeSO3, 10 HEPES, 4 Mg-ATP, 0.3 Na-GTP, 8 Na2-Phosphocreatine, 3.3 QX-314 (pH was adjusted to 7.35 with CsOH). In some experiments, a biocytin (4 mg/mL) and Alexa488/586 (50 μM) were further added for morphological studies. Recordings and hardware control were performed with Wavesurfer (Janelia Farm). Signals were amplified and Bessel filtered at 4 kHz with an amplifier (MultiClamp 700B, Molecular Device), then sampled at 10 kHz with a data acquisition board (USB-6343, National Instrument). Recording with access resistance change of > 20% from the baseline (∼30 MΩ or less) were discarded. Liquid junction potential was not corrected. All recordings were performed at ∼34 °C maintained with the inline heating system (TC-324C, Warner instrument).

To stimulated channelrhodopsin-2, a blue LED was delivered at a short pulse (5-msec) through a 4x objective lens (0.16 NA, UPlanSApo, Olympus). The light power (measured at the focal plane) was 20 mW for the recordings in normal ACSF and 40 mW for the recordings in the presence of tetrodotoxin and 4-aminopyridine. Photo-stimulation was repeated 3 to 6 times (at intervals of 10 or 20 sec) to obtain an average response trace. Photo-evoked responses less than 2 standard deviations of the baseline were considered as a “zero” response. The mean amplitudes indicate mean amplitudes between LED stimulus-onset to 50 ms of average responses.

Drugs used for ex vivo electrophysiology were purchased from Merck, TOCRIS, and Hellobio.

Quantification of presynaptic neurons

Mice were deeply anaesthetized with a ketamine (120 mg/kg) and xylazine (24 mg/kg) mixture and were transcardially perfused with 4% paraformaldehyde (PFA). The brains were extracted and immersed in 4% PFA overnight followed by 20 and 30% sucrose solution for a cryoprotection. The brains were subsequently embedded in O.C.T compound (Tissue TEK, Sakura) in a mold and frozen on the dry ice. Coronal slices (100 µm) of entire brain were cut with Cryostat (CM3050S, Leica) and were washed with phosphate-buffered saline (PBS) before DAPI staining. All sections were mounted onto glass slides with a coverslip (thickness:0.13-0.17 mm, Hounisen) and mounting media (DAKO, S3023, Agilent Technologies).

The injection sites and its specificity were verified by registering the images into Allen Brain atlas using QuickNII and VisuAlign (57).

For the visualization and quantification of presynaptic neuron distribution in a 3D mouse brain, epifluorescence images were acquired using a Slidescanner (4231 x 3462 pixels, 10x objective, Olympus VS120) and analyzed with the MATLAB-based program AMaSiNe (34). Specifically, for Fig. 1, coronal sections between AP +3 mm and -4 mm were registered in AMaSiNe, which semi-automatically detects fluorescently labeled presynaptic cells across all sections. To control for transfection variability (which may arise from slight difference in injection volume or site), we normalized the number of presynaptic neurons in the specified brain areas (shown in Fig. 1H) by the total number of presynaptic neurons in those regions.

For the monosynaptic rabies tracing study in Fig. 6, the brain areas containing over 1% presynaptic cells of total presynaptic cells were proceeded further analysis. Fraction of presynaptic cells in the different areas were calculated by dividing a number of presynaptic cells in each brain areas by the average number of presynaptic neurons in all brain areas.

For the study in Fig. 2, the sections were immunostained with NeuN before imaging. Here, sections were immersed with 5% normal goat serum and 0.2% Triton X for 30 mins at room temperature before incubation with NeuN primary antibody (anti-rabbit, 1:1000) (ABN78, Milipore) at 4 °C for overnight. The sections were then washed with PBS and were further incubated with secondary antibody (goat anti-rabbit conjugated with Alexa 488 or 568, 1:500) (A11008 or A11036, Thermo Fisher) for 1 hour at room temperature before mounting. A single plane image of CLA complex was acquired with confocal microscope (10x objective, 1024 x 1024 pixels, Zeiss). 2-4 images, range of AP: +1.4 to -0.2, were collected per brain. Fraction of double-labelled neurons (with different retrograde tracers) were calculated for each brain by dividing a total number of double-labeled neurons by the total number of vCA1-projecting neurons (labeled with one color).

Quantification of fluorescence signal from axons

For Fig. 3 and S3, horizontal section (100 µm) images were taken with Sliderscanner (10x lens, pixel: 2560 x 3072). All images from the same brain were taken with the same light intensity and detection setting. After background subtraction, GFP signals in ventral, intermediate, and dorsal CA1 column were normalized by the highest signal in ventral CA1 in the same brain.

For Fig. 2, coronal sections (100 µm) were immunostained with GFP antibody before imaging. Here, sections were washed with PBS, the sections were incubated in blocking solution containing 5% normal goat serum and 0.2% Triton X for 30 mins at room temperature before incubation with GFP primary antibody (anti-rabbit, 1:1000) (NB600-308, MNovus Biologicals) in blocking solution at 4 °C for overnight. The sections were then washed with PBS and were further incubated with secondary antibody (goat anti-rabbit conjugated with Alexa 488, 1:300) (A11008, Thermo Fisher) for 1 hour at room temperature before mounting. Images were taken with Slidescanner (10x, pixel: 3072 x 3584). Signal per pixel in ROI was calculated for each projection site after background subtraction.

Social memory test with chemogenetic inhibition of ENvCA1-proj. neurons

Social memory test is based on previous articles with slight modifications (58), (59). All tested mice were habituated to the handling for 3 days. The water-soluble CNO purchased from Hellobio was dissolved by saline (final concentration was 0.1mg/ml). Prior to the test day, we divided subject mice into two groups. On the 1st test day (session 1), one group received chlozapine-N-oxide (CNO, 1 mg/kg, i.p.) and another group received a vehicle. After 30 mins, the subject mice were placed in a three-chamber box for 5 mins (habituation), then the same box with two pencil chambers containing the same object in opposite corners for 5 mins (pretest). After 5 mins break in their home cage, the subject mice were again placed in the same box, but this time, an object in one pencil chamber was replaced with different object and an object in another pencil chamber was replaced with a conspecific (3-4 weeks old male, habituated to the handling and pencil chamber for 3 days prior to the session 1). This “sociability test” lasted for 10 mins and was followed by social discrimination test after 30 mins break in the home cage. The social discrimination test lasted for 5 mins and consisted of the same setting as the sociability test except an object in a pencil chamber, which was replaced with a new conspecific (i.e., a novel mice). In the following day, CNO and saline treatments were swapped between the groups, and all mice went through the same series of tests (session 2) for the within-subject comparison of the CNO effect. The position of the pencil chamber including an object or a conspecific were counterbalanced. The tests were performed under 6-10 lx room light and video was recorded at 25 frames/sec with GigE camera (Basler ace acA1300-60gm, Basler) placed at the ceiling.

Trace fear conditioning

After one week of social memory test, subject mice were examined for their tone-associated fear memory and extinction. One group received CNO while another group received vehicle 45 mins before the test. In conditioning day, mice were placed in a box with a metal grid floor and ethanol odor (context A) and after 3 mins were delivered with 20-sec pure tone (2 kHz, 20 dB) followed by 18-sec time gap (trace) and 2-sec footshock (0.5 mA). The exposure to the same tone-trace-shock sequence was repeated 2 more times with inter-trial interval of 2 mins. In the following day, mice were re-exposed to the context A for 3 mins to assess their fear memory to the context. From day 3, mice were placed in a new context, the context B (smooth floor, breach odor), and went through the same protocol as in conditioning day but without shocks. The process was repeated over 5 days to assess the extinction of the fear memory to the tone and trace. All test was performed in dark (but infrared light was on), and behavior was video recorded at 25 frames/sec with infrared GigE camera (Basler ace acA1300-60gm, Basler).

Novel object recognition test

The novel object recognition test is based on the previous article with slight modifications (58). The subject mice were divided into two groups and CNO treatment was done as same as social memory test. 40 mins after CNO/vehicle i.p. injection, the subject mice were placed in an open field box (50 x 50 cm) for 5 mins (habituation). In following familiarization session, two same shape objects are placed in the open field box, and then the subject mice explored for 10 mins. After 1 hour break in their home cage, the subject mice were again placed in the same box, but one object is replaced by a novel object. The subject mice are allowed to explore the box for 5 min in discrimination session.

Social memory test and novel object recognition test with in vivo fiber photometry

Optic probe implanted mice were habituated handling for 3 days prior the test. In test day, the subject mouse attached with patch code were placed in an open field box (50 x 50 cm) for 5 min (habituation). After habituation, the mouse was place in a home cage for resting and waiting following session. For pretest, pencil chambers containing the same objects were placed at the corners of the open field box. Pretest was tested for 5 min. For sociability test, a novel mouse (4 weeks old, male) or a novel object were placed in pencil chambers. Sociability test took 10 min. After sociability test, the subject mouse return to a home cage and rest for 20 min. After 20 min rest, the subject mouse was tested discrimination session for 5 min. In discrimination test, a novel mouse (4 weeks old, male) or a familiar mouse that was used as the stimulus mouse in sociability test were placed in pencil chambers. Both in sociability and discrimination test, positions of pencil chambers were counterbalanced between subject mice. The tests were performed under 6-10 lux room light and video was recorded at 60 frames/sec with GigE camera (Basler ace acA1300-60gm, Basler) placed at the ceiling. Video acquisitions were controlled by Bonsai. Simultaneous start of video recording and calcium signal recording was controlled by Raspberry Pi.

After one week of social memory test, subject mice were examined for novel object recognition test. The procedure, the arenas and objects were used the same as in the novel object recognition test with chemogenetic inhibition of ENvCA1-proj. neurons.

Behavior data acquisition and analysis

Ethovision (version 16 or 17, Noldus) was used for the hardware control and data acquisition and initial quantification for the behavior. Behavioral variables, including movement trajectory, times spent in predefined areas, total distance traveled, travelling velocity, and freezing were extracted and further analyzed/nested with custom functions in the Matlab (R2020b). For social memory test, interaction zones were defined as an area between the edge of pencil chamber and 3 cm away from the edge. When the nose of the subject mouse was in the zone, the time was counted as interaction time. Sociability index was calculated by dividing total interaction time with mouse by total interaction time with mouse and object. The discrimination index was calculated by dividing total interaction time with a novel mouse by total interaction time with a familiar mouse and a novel mouse. Data from mice were excluded if discrimination index was less than 0.5 during saline treatment. For trace fear conditioning test, mice were considered “freezing” if pixel change between frames was less than 0.1 % for more than 1 sec.

Deeplabcut (ver.2.3) was used for tracking of annotated body points of the subject mice connected with patch code for calcium imaging of ENvCA1-proj. neurons during social memory test and novel object recognition test. Nose points were used for further analysis using Matlab (R2020b). To define interaction zone, reference points was set at the center of pencil chamber. The areas that distance between nose point and reference point was less than 100 pixel for social memory test and 50 pixel for novel object recognition test were dedicated to interaction zone.

Fiber photometry and analysis

Population activity of ENvCA1-proj. neurons were recorded using fiber photometry (Doric Lenses). GCaMP8s was excited with 470 nm light at 20-30 μW measured at the tip (isosbestic point used was 415 nm (15 μW)). Calcium-dependent and independent signals were collected using lock-in mode. Signals were detected at 10x gain. Recorded signals were butterworth filtered (low-pass at 40Hz) and calcium signals (in z-score) were extracted using Doric neuroscience studio V6 (Doric Lenses). Signals were further down sampled to 60 Hz using Matlab (R2020b) signal analyzer to match the behavioral sampling data (60 fps). Z-score > 2.56 (alpha = 0.01) was determined to be a calcium event. For heat map of cumulative time and calcium events, spatial binning was 50 pixels. To make correlation map (Fig. S7 and S8A), correlations between column of cumulative time heat map and cumulative calcium event heat map were tested using Matlab corr function.

Statistical analysis

Data and statistical analysis were performed with MATLAB R2020b (Mathworks). Unless otherwise noted, group data was presented as median, and first and third quartiles are shown for dispersion. Statistics were performed using a non-parametric test (signed-rank test for paired data and rank-sum test for unpaired data). One-way ANOVA (post hoc test: Tukey-Kramer test or Kruskal-Wallis test) were applied for multi-group comparison. Two-way ANOVA were performed for comparison of multi-factors. Pearson correlation coefficient was applied for correlation analysis. Significance was defined as *p < 0.05, **p < 0.01, ***p < 0.001. Outlier was detected by Grubbe test (GraphPad, significant level: 0.05). Sample number were indicated as “n”, animal number were indicated as “N”.

eLife Assessment

This important study offers insights into the function and connectivity patterns of a relatively unknown afferent input from the endopiriform to the CA1 subfield of the ventral hippocampus, suggesting a neural mechanism that suppresses the processing of familiar stimuli in favor of detecting memory guided novelty. The strength of evidence is solid, with careful anatomical and electrophysiological circuit characterization. The work will be of broad interest to researchers studying the neural circuitry of behavior.

Reviewer #1 (Public review):

Summary:

The anatomical connectivity of the claustrum and the role of its output projections has, thus far, not been studied in detail. The aim of this study was to map the outputs of the endopiriform (EN) region of the claustrum complex, and understand their functional role. Here the authors have combined sophisticated intersectional viral tracing techniques, and ex vivo electrophysiology to map the neural circuitry of EN outputs to vCA1, and shown that optogenetic inhibition of the EN→vCA1 projection impairs both social and object recognition memory. Interestingly the authors find that the EN neurons target inhibitory interneurons providing a mechanism for feedforward inhibition of vCA1.

Strengths:

The strength of this study was the application of a multilevel analysis approach combining a number of state-of-the-art techniques to dissect the contribution of the EN→vCA1 to memory function.

In addition the authors conducted behavioural analysis of locomotor activity, anxiety and fear memory, and complemented the analysis of discrimination with more detailed description of the patterns of exploratory behaviour.

Reviewer #2 (Public review):

Summary:

Yamawaki et al., conducted a series of neuroanatomical tracing and whole cell recording experiments to elucidate and characterise a relatively unknown pathway between the endopiriform (EN) and CA1 of the ventral hippocampus (vCA1) and to assess its functional role in social and object recognition using fibre photometry and dual vector chemogenetics. The main findings were that the EN sends robust projections to the vCA1 that collateralise to the prefrontal cortex, lateral entorhinal cortex and piriform cortex, and these EN projection neurons terminate in the stratum lacunosum-moleculare (SLM) layer of distal vCA1, synapsing onto GABAergic neurons that span across the Pyramidal-Stratum Radiatum (SR) and SR-SML borders. It was also demonstrated that EN input disynaptically inhibits vCA1 pyramidal neurons. vCA1 projecting EN neurons receive afferent input from piriform cortex, and from within EN. Finally, fibre photometry experiments revealed that vCA1 projecting EN neurons are most active when mice explore novel objects or conspecifics, and pathway-specific chemogenetic inhibition led to an impairment in the ability to discriminate between novel vs. familiar objects and conspecifics.

Revision 1:

The authors have addressed most of my concerns, but a few weaknesses remain :

(1) I expected to see the addition of raw interaction times with objects and conspecifics for each phase of social testing (pre-test, sociability test, social discrimination), as per my comment on including raw data. However, the authors only provided total distance traveled and velocity, and total interaction time in Figure S9, which is less informative.

(2) The authors observed increased activity in vCA1-projecting EN neurons tracking with the preferred object during the pre-test (object-object exploration) phase of the social tests, and the summary schematic (Figure 9A) depicts animals as showing a preference for one object over the other (although they are identical) in both the social and object recognition tests. However, in the chemogenetic experiment, the data (Fig S9B) indicate that animals did not show this preference for one object over another, making the expected baseline for this task unclear. This also raises an important question of whether the lack of effect from chemogenetic inhibition of vCA1-projecting EN neurons could be attributed to the absence of this baseline preference.

Additionally, the finding that vCA1-projecting EN activity is associated with the preferred object exploration appears to counter the authors' argument that novelty engages this circuit (since both objects are novel in this instance). This discrepancy warrants further discussion.

Author response:

The following is the authors’ response to the previous reviews.

Reviewer #2 (Public review):

Summary:

Yamawaki et al., conducted a series of neuroanatomical tracing and whole cell recording experiments to elucidate and characterise a relatively unknown pathway between the endopiriform (EN) and CA1 of the ventral hippocampus (vCA1) and to assess its functional role in social and object recognition using fibre photometry and dual vector chemogenetics. The main findings were that the EN sends robust projections to the vCA1 that collateralise to the prefrontal cortex, lateral entorhinal cortex and piriform cortex, and these EN projection neurons terminate in the stratum lacunosum-moleculare (SLM) layer of distal vCA1, synapsing onto GABAergic neurons that span across the Pyramidal-Stratum Radiatum (SR) and SR-SML borders. It was also demonstrated that EN input disynaptically inhibits vCA1 pyramidal neurons. vCA1 projecting EN neurons receive afferent input from piriform cortex, and from within EN. Finally, fibre photometry experiments revealed that vCA1 projecting EN neurons are most active when mice explore novel objects or conspecifics, and pathway-specific chemogenetic inhibition led to an impairment in the ability to discriminate between novel vs. familiar objects and conspecifics.

The authors have addressed most of my concerns, but a few weaknesses remain :

(1) I expected to see the addition of raw interaction times with objects and conspecifics for each phase of social testing (pre-test, sociability test, social discrimination), as per my comment on including raw data. However, the authors only provided total distance traveled and velocity, and total interaction time in Figure S9, which is less informative.

We apologies for missing the request. We have added the raw interaction times in Fig. S9G.

(2) The authors observed increased activity in vCA1-projecting EN neurons tracking with the preferred object during the pre-test (object-object exploration) phase of the social tests, and the summary schematic (Figure 9A) depicts animals as showing a preference for one object over the other (although they are identical) in both the social and object recognition tests. However, in the chemogenetic experiment, the data (Fig S9B) indicate that animals did not show this preference for one object over another, making the expected baseline for this task unclear. This also raises an important question of whether the lack of effect from chemogenetic inhibition of vCA1-projecting EN neurons could be attributed to the absence of this baseline preference.

We appreciate the comments. In Fig. S9B, although the group median at baseline (pretest) showed no preference for one object, individual subjects displayed a preference for one object (i.e., each data point deviated positively or negatively from 0.5) in saline condition. Therefore, we do not think that a lack of baseline preference accounts for the absence of the inhibition effect in the pretest.

Additionally, the finding that vCA1-projecting EN activity is associated with the preferred object exploration appears to counter the authors' argument that novelty engages this circuit (since both objects are novel in this instance). This discrepancy warrants further discussion.

This is an interesting point. One possibility is that during the pretest, EN activity simply "reports" or "represents" the interaction time without driving exploratory preference. This aligns with our DREADD experiment data, which show that inhibition of EN neurons produced no overall behavioral effect. Innate exploratory behavior has been attributed to various circuits, including the medial preoptic area → PAG circuit (Ryoo et al., 2021, Front. Neuro.) and the Septal → VTA circuit (Mocellin et al., 2024, Neuron). We found no direct projection from these areas to EN (Fig. 6), but such connections could be established di- or polysynaptically. Moreover, these circuits could be driven by common inputs, such as the locus coeruleus or the cholinergic system for arousal, with only specific downstream targets, excluding EN, playing a key role in driving innate exploration and preference.

We have inserted the following sentence in discussion (line 253-255):

“The correlation of ENvCA1-proj. activity with novel object preference in the pretest nevertheless suggests that these neurons 'represent' the innate preference without driving it.”

Recommendations for the authors:

Reviewer #2 (Recommendations for the authors):

Line 209: Please remove the reference to neural activity 'predicting' behavior, as correlation analysis does not imply predictive power.

We now have changed the phrase to “Although ENvCA1-proj. activity was correlated with the behavior…”

Line 236: It is unclear what is meant by: 'This circuit motif may predict the predominant role of ENvCA1-proj. neurons in social recognition memory'

We have changed the sentence to the following for the clarity:

“Since social odor information is crucial for discriminating conspecifics in rodents, this circuit motif may predict the predominant role of ENvCA1-proj. neurons in social recognition memory, given that social odor can engage multiple olfactory pathways innervating the piriform cortex.”

Fig 7 title: insert 'with' after correlates: 'Activity of ENvCA1-proj. neurons correlates social/object discrimination performance'

Corrected.

Fig S1 title: 'Projecing' typo.

Corrected.

Fig S8: Please rephrase for clarity: 'In pretest, the object was aligned by longer interaction time (preferred object is plotted in right side)'

We now have rephrased the sentence to:

“In the pretest plot, the object that the mice interacted with more is placed on the right side.”

References:

A septal-ventral tegmental area circuit drives exploratory behavior. Mocellin, Petra et al. Neuron, Volume 112, Issue 6, 1020-1032.e7

An inhibitory medial preoptic circuit mediates innate exploration. Ryoo, Jia et al. Front. Neurosci., 23 August 221. Volume 15- 2021

Version 3

Abstract

Abstract

The claustrum complex is viewed as fundamental for higher order cognition; however, the circuit organization and function of its neuroanatomical subregions are not well understood. We demonstrated that some of the key roles of the CLA complex can be attributed to the connectivity and function of a small group of neurons in its ventral subregion, the endopiriform (EN). We identified a subpopulation of EN neurons by their projection to the ventral CA1 (ENvCA1-proj. neurons), embedded in recurrent circuits with other EN neurons and the piriform cortex. Although the ENvCA1-proj. neuron activity was biased toward novelty across stimulus categories, their chemogenetic inhibition selectively disrupted the memory-guided but not innate responses of mice to novelty. Based on our functional connectivity analysis, we suggest that ENvCA1-proj. neurons serve as an essential node for recognition memory through recurrent circuits mediating sustained attention to novelty, and through feed forward inhibition of distal vCA1 neurons shifting memory-guided behavior from familiarity to novelty.

Introduction

The claustrum (CLA) complex, an evolutionarily conserved brain region found across mammalian species, reptiles, and birds, is hypothesized to serve as a node for establishing higher order cognitive functions by coordinating neuronal activity on a global scale (1–3). This view is based on its extensive connections with many cortical and subcortical areas (4, 5). Some of these functions include sensory perception and attention, which may affect memory processing, including working memory, associative memory, and recognition memory (6–8). Consistent with this view, abnormalities of the CLA complex is found in major cognitive disorders, including Alzheimer’s disease, schizophrenia, and attention-deficit/hyperactivity disorders (9–11). However, microcircuitry and functional segregation of the individual constituents of the CLA complex have remained unclear.

The cytoarchitecture and genetic expression broadly divide the CLA complex into dorsal and ventral parts, with the dorsal part typically referred to as “CLA” and ventral part having its own nomenclature: “endopiriform (EN)” in rodents (12, 13). Current evidence indicates that the CLA forms reciprocal connections with dorsal cortices containing sensory, motor and association areas generating diverse physiological effects depending on the cortical targets and their activity patterns (4, 5). In contrast, EN primarily provides inputs to the piriform cortex and limbic systems (14, 15), suggesting its distinct circuit organization and role in bridging olfactory information processed by the piriform cortex and memory processed by the limbic systems (16, 17).

One of distinct targets of EN, the ventral CA1 (vCA1), is known to coordinate in a range of behaviors related to exploration and recognition memory, and these functions are thought to be regulated by its afferent inputs in a domain-specific manner (18–26). We, therefore, hypothesized that EN is a key node for establishment of recognition memory. To address this, we set out to characterize circuit and function of EN neurons defined by their projection to vCA1 (EN vCA1-proj. neurons) using genetic tools and mouse models of social and object recognition memory.

We found EN vCA1-proj. neurons innervated multiple components of the limbic system except amygdala and prefrontal cortex and produced potent feedforward inhibitory control over vCA1 pyramidal neurons. During recognition memory test, activity of EN vCA1-proj. neurons were condensed around conspecifics or objects where mice spent most time on. However, disruption of EN vCA1-proj. activity only impaired memory-guided exploration of novel stimuli without affecting innate exploration induced by novelty. These findings demonstrate that EN subserves some of the key functions required for recognition memory governed by specific limbic system.

Results

Endopiriform represents a major afferent of the ventral CA1

To assess the significance of EN as a vCA1 afferents, we injected a retrograde tracer (60 nl) into the vCA1 of mice and compared the number of the retrogradely labeled neurons in the EN and other brain regions (Fig. 1A, B, S1A, Methods). We found labeled neurons in multiple areas including the septum, entorhinal cortex, and basolateral amygdala (Fig. 1C-F and Fig. S1B) (27, 28). In addition to these well-known vCA1 afferents, prominent labeling was observed in the area defined as EN in Allen Brain Atlas (Fig. 1G, Methods). The number of presynaptic neurons in EN was ∼10 fold lower than entorhinal cortex (data not shown). However, the normalized count of presynaptic neurons in EN was significantly greater than basolateral amygdala, lateral septum, and medial septum (Fig. 1H).

Endopiriform projects to the ventral CA1

(A) Schematic diagram of a retrograde tracer injection performed. CAV-2-Cre or CTB647 was used in combination with Ai14 or wild-type mice, respectively. (B) An example fluorescent image of injection site. (C) Distribution of retrogradely labeled vCA1-projecting neurons in entire brain as a result of injection in panel B. (D-G) Fluorescent images showing retrogradely labeled vCA1-projecitng neurons in septum (SP), entorhinal cortex (ENT), basolateral amygdala (BLA), and endopiriform (EN). Str: Striatum, CLA: Claustrum. (H) Box plots showing relative number of vCA1-projecting neurons in different brain areas. Cell number in each brain area were normalized by summed cell number of EN, BLA, SP (lateral), and SP (medial). Cell numbers in each brain areas of individual brains are presented in Table 1. Median value for EN, BLA, SP (lateral), and SP (medial) is 0.45, 0.24, 0.08, and 0.21, respectively. Statistics: EN vs BLA, SP (lateral), or SP (medial), p < 0.001 for all; BLA vs SP (lateral) or SP (medial): p = 0.002 and 0.694, respectively; SP (lateral) vs SP (medial) p = 0.023. One-way ANOVA, repeated comparison with Tukey-Kramer test. N = 5. (I) An example fluorescent image of horizontal section showing retrogradely labeled vCA1-projecting neurons in EN. A: anterior, P: posterior. (J) Top: Example map showing distribution of vCA1-projecting neurons in entire EN in one mouse. Bottom left: A plot showing a quantified distribution at bin size of 1 mm. Each bin was normalized to total number of labeled neurons in EN. Data from each mouse is shown in grey. Median is indicated in black. Bottom right: Box plots comparing a number of vCA1-projecting neurons in anterior and posterior EN. Data were normalized to a total number of labeled neurons in EN. Median value for anterior vs posterior was 0.39 vs 0.54. Statistics: p = 0.008, Rank-sum.

The vCA1-projecting neurons in the EN were spread along antero-posterior axis with two peak densities, one in anterior and another in posterior to the bregma (Fig. 1I, J). Number of labeled neurons was greater in posterior EN (Fig. 1J). In contrast, the same analysis along dorso-ventral axis showed a single dominant peak at the depth consistent with the location of EN (Fig. S1C). To further confirm the location of these vCA1-projecting neurons, we injected a retrograde tracer (60 nl for each) in one color into vCA1 and a retrograde tracer in another color into a cortical area known to receive projections from EN or “CLA” representing a dorsal part of CLA complex (Fig. 2A). We targeted the prefrontal cortex to label EN, and the anterior cingulate cortex, the motor cortex, and the dorsomedial entorhinal cortex to label CLA (Fig. 2C-F) (29–32).

Axon branching of vCA1-projecting EN neurons

(A) Schematic diagram of injection performed and two hypothesized outcomes. Yellow color indicates double-labeled neurons. PFC: Prefrontal cortex, ACC: Anterior cingulate cortex, MC: Motor cortex, ENT(dm): Dorsomedial entorhinal cortex, CLA: claustrum (B-F) Example fluorescent images of injection site for each region indicated. (G-J) Example fluorescent images of retrogradely labeled neurons in CLA complex. vCA1-projecting neurons were colored in red and cortex-projecting neurons were colored in green. (K) Box plots showing percentage of double-labeled vCA1-projecting neurons for each cortical injection. Medial value for PFC, ACC, MC, and ENT(dm) was 28.6, 4.6, 0, and 1.2, respectively. Statistics: PFC vs ACC, MC or ENT(dm), p = 0.405, 0.015 and 0.047, respectively. ACC vs MC or ENT (dm), p = 0.422 and 0.64, respectively. MC vs ENT (dm), p = 0.993. One-way ANOVA with post-hoc Kruskal-Wallis test. Slice and animal number: 16 and 4 for PFC, 20 and 5 for ACC, 10 and 3 for MC, and 18 and 4 for ENT(dm). (L) Schematic diagram of projection pattern found in the experiment shown in A. (M) Schematic diagram of the intersectional approach performed to label vCA1-projecting EN neurons. (N-O) Example fluorescent images showing vCA1-projectig axons in different brain areas. All images were from the same mouse. (P) Magnified images showing vCA1-projecting axons in cortical layers. Magnified areas are indicated in O and fig. S2A-C as dashed squares. PIR: Piriform cortex

We found there is a major spatial overlap in labeling of vCA1-projecting neurons with prefrontal cortex-projecting neurons, but not with other cortical projection neurons (Fig. 2G-J). Moreover, around 30% of vCA1-projecting neurons were found to be double-labelled when a cortical injection was made into the prefrontal cortex, suggesting some vCA1-projecting neurons send axons to this cortical area (Fig. 2K, L).

To further determine the cortical and subcortical innervation pattern of vCA1-projecting EN neurons, we used an intersectional approach to label their somata and axons with eGFP. CAV2-Cre (60 nl) was injected into the vCA1 and flox-eGFP (100 nl) was injected into the EN (Fig. 2M). Apart from vCA1, axons from vCA1-projecting neurons were found in various cortical areas including prefrontal cortex, lateral entorhinal cortex, and piriform cortex (Fig. 2N-P and Fig. S2A-C). Projection to prefrontal cortex was much sparser relative to that of vCA1, as expected based on the retrograde labeling data (Fig. 2K). The connectivity of vCA1-projecting EN neurons to the amygdala, which represents the major component of limbic systems, was also sparse relative to vCA1 (Fig. S2C), indicating vCA1 is the main target of these EN neurons in this system. Within the hippocampus, very sparse EN axons were observed in vCA3 and dentate gyrus (Fig. S2D).

Taken together, these findings indicate EN represents a major afferent of vCA1. Moreover, EN neurons projecting their axons to vCA1 also send strong collaterals to lateral entorhinal cortex and piriform cortex, but relatively weak to the prefrontal cortex or other limbic structures. Since our subsequent study focused on EN neurons defined by their projection to vCA1, we refer to them as ENvCA1-proj. neurons.

Projection pattern and synaptic connectivity of EN axons in vCA1

We next investigated projection pattern of EN axons in hippocampus by injecting an anterograde tracer (AAV-GFP, 60 nl) into the EN (Fig. 3A, B). Analysis of different parts of hippocampal section indicated EN axons were mostly confined in SLM layer of distal part of vCA1 (Fig. 3C, D). Projection to intermediate or dorsal CA1 was limited or undetectable (Fig. S3A-C). Since SLM mainly consists of GABAergic neurons (33), we hypothesized that EN axons form synapses with this cell type. To test this, we first applied monosynaptic rabies tracing technique to Vgat-Cre mice (Fig. 4A, Methods). The distribution of starter cells (identified by GFP and mCherry co-labeling) was confirmed to be within the ventral-intermediate region of hippocampus using AMaSiNe (Fig. 4B, S4A) (34). In these mice, the presynaptic neurons were consistently observed in EN in addition to other expected areas (e.g., septum and entorhinal cortex) (Fig. 4C and D, S4B-F) (35, 36). The density of the presynaptic neurons in EN was greater at the posterior region to the bregma (Fig. 4D), consistent with our earlier observation (Fig. 1J).

EN projection pattern in vCA1

(A) Schematic diagram of injection performed. (B) An example fluorescent image of horizontal brain section containing injection site in EN. (C) An example fluorescent image of ventral hippocampal slice with EN axons resulted from the injection in panel B. Dashed lines indicate the distal and proximal vCA1 regions where GFP signals were measured for the plot in panel D. (D) Normalized GFP signals across the layer. All signals were normalized by maximum GFP signal in the distal subregion. Median of area under the curve: distal; 11.12, proximal; 5.99. distal vs proximal, p = 0.016. Wilcoxon signed-rank test. n = 8 slices (N = 4 mice). SR: stratum radiatum, Py: pyramidal layer.

ENvCA1-proj. neurons target vCA1 interneurons

(A) Schematic diagram of injection performed for monosynaptic rabies tracing. Flox-TVA-GFP (60 nl) was injected into the vCA1. 4 weeks later, Rabies-mCherry (60 nl) was injected. (B) Location of starter cells in the hippocampus. Data from 5 brains (each brain represented by different color) were overlaid. (C) Example images showing presynaptic cells in EN (white arrow). (D) Top: Location of presynaptic EN neurons. Data from 5 brains were overlaid (different brains were coded by different colors). Bottom: Plots indicate distribution of presynaptic cells in EN along AP axis. Cell counts in a given bin (1 mm) were normalized by total cell number in EN. Thin gray lines indicate individual data. A thick black line indicates median. Inset showing plot comparing cell number in anterior and posterior location as in Fig. 1J. Median: anterior vs posterior, 0.25 vs 0.72, p = 0.031; Wilcoxon signed-rank test. (E) Top: Schematic diagram of injection performed. Bottom: An example fluorescent image of injection site in EN. (F) Schematic diagram of recordings performed. Recordings were performed from identified GABAergic neurons sequentially in a random order. (G) An example confocal image of an acute vCA1 slice used for recordings. Recorded GABAergic neurons were filled with biocytin and post-labeled with Alexa488. White dashed lines indicate border of Py-SR and SR-SLM. (H) Median traces of photo-evoked EPSCs recorded from neurons at different laminar position. (I) Connection probability (CP) of EN axons and recorded neurons at different laminar positions. Numbers in bars indicate number of response positive neurons /total recorded neurons. (J) Left: Peak amplitudes of evoked EPSCs (normalized to total input from recorded slice). Right: the same as left but for mean amplitude. Gray circles indicate each neuron. Black lines indicate median. Median (peak amp): Py-SR vs SR, SR-SLM or SLM: 1.1 vs 1.0 vs 1.1 vs 0.0. [Py-SR vs SR, SR/SLM vs SLM], p = 0.999, 0.988, 0.076; [SR vs SR-SLM or SLM], p = 0.972, 0.111; [SR-SLM vs SLM], p = 0.039; one-way ANOVA, repeated comparison with Tukey-Kramer test. Median (mean amp): Py-SR vs SR, SR-SLM, or SLM: 1.0 vs 0.8 vs 1.4 vs 0.0. [Py-SR vs SR, SR-SLM or SLM], p = 0.991, 0.999, and 0.141; [SR vs SR-SLM or SLM], p = 0.978, and 0.251; [SR-SLM vs SLM], p = 0.114; one-way ANOVA, repeated comparison with Tukey-Kramer test. N = 5 mice.

To determine how EN inputs are integrated into the local circuit in vCA1, we systematically assessed the connections between EN axons and GABAergic neurons in different layers of distal vCA1. We expressed channelrhodopsin-2 (ChR2) into the EN axons by injecting AAV-ChR2-mCherry (100 nl) into EN and performed whole-cell recording from GABAergic neurons in acute vCA1 slices in the presence of TTX (1 µM) and 4-AP (100 µM) to isolate monosynaptic connection (37). To ensure the recording from GABAergic neurons, we used Vgat::Ai14 mice in which all GABAergic neurons are labeled with tdTomato (Fig. 4E-G). We recorded from 3 to 4 GABAergic neurons located in different laminar position in distal vCA1 (Py-SR border, SR, SR-SLM border, or SLM) in the same slice to construct a “laminar profile” of EN◊vCA1 inputs (Fig. 4F and G). Whole-field LED stimulation evoked excitatory postsynaptic current (EPSC) in most GABAergic neurons recorded. However, contrary to the expectation, connection probability and strength were lowest in neurons in SLM compared to neurons in other layers (Fig. 4H-J). Taken together, these results indicate EN axons preferentially innervate the GABAergic neurons spanning across the Py-SR border to SR-SLM border in distal vCA1.

EN axons produce feedforward inhibition onto vCA1 pyramidal neurons

SLM also contains the apical tuft dendrite of pyramidal neurons, which could be a target of EN axons in addition to GABAergic neurons. To test this, we expressed ChR2 in EN axons but made whole-cell recording from pyramidal neurons in distal vCA1 in the presence of TTX and 4-AP. Since CA1 pyramidal neurons can be differentiated into superficial and deep layer neurons with distinct anatomy and function (38), we recorded from neurons in each layer as well as nearby GABAergic neurons at the Py-SR and SR-SLM borders for the comparison (Fig. 5A). We found monosynaptic input to superficial and deep layer pyramidal neurons was similar, and monosynaptic input to pyramidal neurons (superficial and deep pooled) and GABAergic neurons was also comparable (Fig. S5).

EN disynaptically inhibits vCA1 pyramidal neurons

(A) A confocal image of recorded pyramidal neurons in vCA1. Recorded pyramidal neurons were filled with biocytin and post-labelled with Alexa488. Ori.: oriens, Super.: superficial layer pyramidal neuron. Deep: deep layer pyramidal neuron. (B) Example traces of photo-evoked EPSCs and IPSCs in superficial (black trace) and deep (gray trace) pyramidal neurons. Red trace: after TTX and 4AP. Blue trace: after gabazine. (C) Scatter plots of peak amplitudes of EPSCs and IPSCs for each neuron. Dashed line indicates a unitary line. Superficial neurons: empty Black circles. Deep neurons: Gray circles. N = 4 mice. (D) A summary diagram of EN◊vCA1 circuit. In vCA1, EN axons innervate GABAergic neurons (IN), which in turn inhibits pyramidal neurons (both superficial and deep).

Since EN axons innervated multiple GABAergic neurons across different layers, we hypothesized they may exert stronger inhibition onto pyramidal neurons through feed-forward inhibition. To test this, we recorded EPSC and inhibitory postsynaptic current (IPSC) from superficial and deep pyramidal neurons using cesium-based internal solution in the brain slices expressing ChR2 in the EN axons (Methods). At the command potential of −70 mV, photo-stimulation of EN axons evoked EPSCs in majority of pyramidal neurons in both layers (7 out of 10 neurons in superficial and 5 out of 10 neurons in deep) (Fig. 5B). At the command potential of +12 mV, the same stimulation evoked outward postsynaptic current in the same neurons, including those without EPSCs (10 out of 10 for superficial and deep) (Fig. 5B). This outward current was abolished by application of TTX and 4-AP or gabazine (10 µM), indicating it is disynaptically driven and mediated by inhibitory GABAA receptor (Fig. 5B). When relative strength of EPSCs and IPSCs were compared for each neuron, the strength of IPSCs overwhelmed the EPSCs in all cases (Fig. 5C). This indicates EN axons disynaptically inhibit pyramidal neurons in vCA1 (Fig. 5D).

ENvCA1-proj. neurons receive inputs from the piriform cortex

We next examined afferents that may drive the EN◊vCA1 circuit. To address this, retrograde CAV2-Cre was injected into vCA1 to express Cre into ENvCA1-proj. neurons, then applied monosynaptic rabies tracing technique to specifically label its presynaptic neurons in wild type mice (Fig. 6A). Starter cells and presynaptic neurons were spatially mapped and quantified using AMaSiNe (Fig. 6B and C, S6). Since presynaptic neurons were sparse in contralateral hemisphere, data from ipsilateral hemisphere was analyzed. Moreover, we applied minimum labeling threshold to determine the brain region for further analysis (Methods). We found the piriform cortex contained highest number of presynaptic neurons followed by EN (Fig. 6D-F). Within piriform cortex, layer 2 neurons were the major source of afferent (Fig. 6G). These data indicate ENvCA1-proj. neurons receive major input from piriform cortex and form a recurrent connection with neurons in EN (Fig. 6H).

ENvCA1-proj. neurons receive inputs from the piriform cortex and within EN

(A) Schematic diagram of performed. After CAV2-Cre (60 nl) injection into the vCA1, flox-TVA-mCherry (100 nl) was injected into the EN. 4 weeks later, Rabies-GFP (100 nl) was injected into the EN. (B) Location of starter cells in EN. Data from 6 brains were overlaid. (C) An example map of presynaptic cells in one brain. Top: Horizontal view. Injections were made in left hemisphere. Bottom: Front view. (D) Number of presynaptic cells in different brain areas (normalized to total presynaptic cell number). Median for AIA, AON, PIR, ENT(lateral), CLA, EN, Pallidum, and SI (in fraction): 0.7, 0.3, 3.8, 0.3, 0.5, 1.7, 0.3, and 0.2. [AON, ENT(lateral), CLA, Pallidum, SI] vs PIR: p = 0.016, 0.008, 0.048, 0.006, and 0.003. EN vs SI, p = 0.043. one-way ANOVA with post-hoc Kruskal-Wallis test. Full list of statistical tests and p values are in Table 2. (E) An example fluorescent image of coronal slice containing EN and PIR and labeled presynaptic cells. (F) Example maps of presynaptic cells in PIR (top) and EN (bottom). (G) Number of presynaptic cells in PIR in different layers (normalized to the total number of labeled neurons in PIR). Median: L1 vs L2 vs L3, 0.11 vs 0.61 vs 0.27. L1 vs L2, p < 0.001; L1 vs L3, p = 0.005; L2 vs L3, p < 0.001; one-way ANOVA, repeated comparison with Tukey-Kramer test. (H) Summary diagram showing input-output circuit of ENvCA1-proj. neurons (red circle).

The activity of ENvCA1-proj. neurons is correlated with the time mice spend in space

The downstream target of EN, vCA1, is implicated in social, odor and object recognition memory (20–22). To determine whether and how the activity of EN neurons is related to this function, we expressed GCaMP8s in ENvCA1-proj. neurons and monitored their activity using fiber photometry while video recording mice performing social recognition memory test consisting of pretest, sociability test, and discrimination test (Fig. 7A, B, Methods). The nose position of the subject mouse was tracked during tests using DeepLabCut (Fig. 7C) (39, 40), and cumulative time and calcium event were mapped onto the arena space (Fig 7C, S7). We found these values were correlated in space across all sessions (i.e., the more time the mice spent in a given space, the more the ENvCA1-proj. activity), including in the open field with no object or in the test arena with conspecifics or objects (Fig. S7, S8A-B).

Activity of ENvCA1-proj. neurons correlates with social/object discrimination performance

(A) Right and Middle: Schematic diagrams of injection and optic cannula implantation performed. CAV2-Cre (60 nl) was injected into the vCA1, and dlox-jGCaMP8s (100 nl) was injected into the EN. Left: An example image showing jGCaMP8s expressing ENvCA1-proj. neurons and a tract of optic cannula. (B) The diagram of social recognition memory test. A test consisted of pretest (5 min), sociability test (10 min), and discrimination test (5 min). (C) Left: An example image of a mouse with nose point marker tracked by DLC. Right: An example map of nose points and calcium events in the arena during sociability test. (D) Nose distance vector (distance between subject’s nose point and a center point of novel/familiar mouse chambers) and simultaneously acquired calcium signals during discrimination test. Gray dashed lines indicate the border of interaction zones. Open red circles indicate the calcium event detected (threshold: Z > 2.58). A part of calcium signal trace was expanded at the bottom for clarity. (E) Left: Correlation between calcium event ratios (calcium event number in the mouse interaction zone/ calcium event number in the object interaction zone) and sociability index. p = 0.29, N = 8 mice. Right: same as in left panel, but for calcium event ratios (calcium event number in the unfamiliar mouse interaction zone/ calcium event number in the familiar mouse interaction zone) and discrimination index. p = 0.003, N = 8. (F) The diagram of novel object recognition memory test. A test consisted of habituation (5 min), familiarization (10 min), and discrimination test (5 min). (G) Same as in (E), but for calcium event number ratios (calcium event number in the unfamiliar object interaction zone/calcium event number in the familiar object interaction zone and discrimination index. p = 0.01. N = 9 mice.

Cumulative calcium events of ENvCA1-proj. neurons became high towards the space around objects or conspecifics when they are present and was correlated with interaction times except for sociability session (Fig. S8C-D). In the sociability session, the cumulative calcium events around a novel conspecific and object were similar despite mice spending more time on conspecifics (Fig. 7E left, S8C). In contrast, in the social discrimination test, both the cumulative time and cumulative ENvCA1-proj. activity were higher for an unfamiliar than familiar conspecific (Fig. 7D-E, S8C). As a result, ENvCA1-proj. activity and behavior had a greater degree of correlation in social discrimination than sociability.

We also tested if a similar correlation occurs in a non-social context by recording ENvCA1-proj. activity and behavior in the object discrimination test (Fig. 7F). The cumulative time and calcium events were significantly correlated across different sessions i.e., familiarization, and object discrimination (Fig. 7G, S8E), largely consistent with the findings in the social discrimination test.

These data indicate that the activity of ENvCA1-proj. neurons is generally correlated with the time the mouse spends in a particular location at the basal level (i.e. open field) or task conditions. However, the degree of correlation appears to vary depending on the session, such that it was highest in the pretest/familiarization and social/object discrimination test, but lowest in the sociability test.

Inhibition of ENvCA1-proj. neurons impairs social/object recognition memory

Although ENvCA1-proj. activity was correlated with the behavior in pretest/familiarization and discrimination test better than sociability test, their causality is unclear. To address this, we inhibited the activity of ENvCA1-proj. neurons using inhibitory DREADD (hM4Di) during social and object discrimination tests (Fig. 8A-C and h, S9A) (41). To specifically express hM4Di in ENvCA1-proj. neurons, CAV-2-Cre (60 nl) and AAV-flox-hM4Di (100 nl for each coordinate) were bilaterally injected into vCA1 and EN, respectively (Fig. 8A-B). hM4Di was replaced with tdTomato for control group. We made a protocol that allows within-subject comparison, such that a mouse goes through a test with chlozapine-N-oxide (CNO, 1 mg/ kg) treatment then the same test with saline treatment the next day, or vice versa (Fig. 8C and H). In social discrimination test, CNO treatment did not affect the pretest or sociability in hM4Di or control group, but specifically impaired social discrimination in hM4Di group (Fig. 8D-G, S9B-G). A similar effect was also observed in novel object discrimination (Fig. 8I, S9D).

Chemogenetic inhibition of ENvCA1-proj. neurons specifically impairs recognition memory

(A) Schematic diagram of injection performed. (B) Left: An example fluorescent image showing expression of hM4Di in ENCA1-proj. neurons. Right: Expansion of areas marked by white rectangle in image on the left. (C) The diagram of social recognition memory test. A session consisted of habituation (5 min), Pretest (5 min), Sociability test (10 min), and discrimination test (5 min). The group treated with a saline in session 1 was treated with CNO (1mg/ kg) in session 2 the following day. (D) Example cumulative times heat map in arena during sociability tests for the same mouse under two different treatments. (E) Plots comparing sociability index for hM4Di and tdTomato mice under two different treatments. Gray empty circles indicate each mouse. Gray thin lines show paired data from one mouse. Thick black bars indicate median. Median sociability index: hM4Di mice (Saline vs CNO), 0.75 vs 0.76, p = 0.391; signed-rank test; tdTomato mice (Saline vs CNO), 0.77 vs 0.75, p = 0.383; signed-rank test. (F) Example cumulative time heat maps in arena during discrimination tests for the same mouse under two different treatments. (G) Comparison of discrimination index for hM4Di and tdTomato mice under two different treatments. Gray empty circles indicate each mouse. Gray thin lines show paired data from one mouse. Thick black bars indicate median. Median discrimination index: hM4Di mice (Saline vs CNO), 0.66 vs 0.57, p = 0.042; signed-rank test; tdTomato mice (Saline vs CNO), 0.65 vs 0.63, p = 0.742; signed-rank test. (H) The diagram of novel object recognition memory test. A session consisted of habituation (5 min), familiarization (10 min), and discrimination test (5 min). CNO treatment was the same as in C. (I) Same as in G, but for novel object discrimination index. Median discrimination index: hM4Di mice (Saline vs CNO), 0.64 vs 0.5, p = 0.003; signed-rank test.

Apart from recognition memory, vCA1 is implicated in anxiety and associative fear memory (42, 43), thus, we tested contribution of ENvCA1-proj. neurons to these functions. The anxiety level as determined by locomotor activity (44) and sociability was not affected by CNO treatment in hM4Di group (Fig. 8E, S9E). Similarly, fear memory as assessed by the freezing response during training (with trace fear conditioning) or recall to context or tone was not affected by CNO treatment in hM4Di group (Fig. S10).

Taken together, these data indicate that ENvCA1-proj. activity is not causally related to innate exploration behavior induced by novelty, or anxiety or fear memory. However, their activity is necessary for mice to discriminate between familiar and unfamiliar conspecific or object, suggesting their major role in general recognition memory.

Discussion

We characterized the circuit and function of EN neurons targeting specific subregions and layers of vCA1. In contrast to other parts of CLA complex, ENvCA1-proj. neurons were reciprocally connected with the piriform cortex representing a major downstream target of the olfactory bulb (45). Since social odor information is crucial for discriminating conspecifics in rodents, this circuit motif may predict the predominant role of ENvCA1-proj. neurons in social recognition memory, given that social odor can engage multiple olfactory pathways innervating the piriform cortex (46). However, we found that these neurons contributed to broader domains of recognition memory. This is supported by the observations that the inhibition of ENvCA1-proj. activity impaired both social and non-social discrimination.

In all phases of recognition memory test, EN vCA1-proj. activity was highly correlated with the time mice spent in given locations and this correlation was biased towards novel stimuli when mice were engaged in their exploration irrespective of the stimuli being social or non-social (Fig. 9A). Taken together, these data suggest that the function of ENvCA1-proj. neurons was primarily related to the detection of and attention to novel stimuli. The recurrent connections within EN, which is another prominent feature of ENvCA1-proj. neurons’ network, potentially support these processes to help establish the recognition memory (47).

The role of ENvCA1-proj. neurons in recognition memory.

(A) Summary of findings from in vivo experiments. (B) The model of ENvCA1-proj. circuit function in recognition memory. During memory-guided exploration, recurrent circuits in ENvCA1-proj. neurons maintain the attentional response to novel stimuli while regulating the degree of response to familiar stimuli in vCA1 via feedforward inhibition.

Attentional processes mediated by ENvCA1-proj. neurons can be considered of particular relevance for memory-guided behavior, given that inhibition of ENvCA1-proj. activity selectively impaired conspecific/object discrimination but not innate exploratory behavior provoked by novelty as seen in pretest, sociability, and familiarization (Fig. 9A). The correlation of ENvCA1-proj. activity with novel object preference in the pretest nevertheless suggests that these neurons ‘represent’ the innate preference without driving it. This functional specialization, likely associated with its unique circuit connectivity to the limbic system, could differentiate EN from CLA.

Studies on EN are generally scarce, and the available evidence of EN projections to vCA1 suggests that these projections are sparse (14). Potential reasons include the difficulty in clearly delineating the EN from CLA complex and non-uniform distribution of ENvCA1-proj. neurons along antero-posterior axis. By leveraging contemporary circuit analysis tools, we demonstrated that the EN◊vCA1 circuit is non-trivial; ENvCA1-proj. neurons were more numerous than commonly studied vCA1 afferents, including those from the septum and amygdala. Moreover, these axons exhibited a potent disynaptic inhibition of distal vCA1 pyramidal neurons, indicating their major functional implications for vCA1 activity.

Our data indicated that although EN axons terminated at the SLM of distal vCA1, they exhibited stronger connections to GABAergic neurons spanning from the Py-SR border to the SR-SLM border than to SLM neurons. The detection of EN inputs from GABAergic neurons at the Py-SR border was surprising, considering the distance between soma position and the EN axons. However, there are several GABAergic cell types below SLM that extend their primary dendrites into SLM (48, 49), likely explaining the observed connectivity pattern. Among these are chandelier cells found at the Py-SR border, which preferentially inhibits the axon initial segments of pyramidal neurons, and are thus well suited to exert the powerful feedforward inhibition of vCA1 pyramidal neurons observed in our recordings (50).

There are several potential circuit effects by which EN inputs could affect vCA1 function. One is the induction of non-linear events in vCA1 pyramidal neurons by acting in concert with other afferents (e.g., lateral entorhinal cortex targeting SLM) (36) to generate a specific output pattern in pyramidal assemblies representing task-relevant stimuli. Alternatively, EN inputs may improve signal to noise ratio of the information conveyed through the lateral entorhinal cortex, CA2 (in the case of social context), and possibly other afferents (e.g., CA3) (51). Another scenario is tuning of vCA1 pyramidal neurons’ response to familiar stimuli through dynamic combination of a weak monosynaptic excitatory input and strong disynaptic inhibitory input. Such mechanism suggests that at a milliseconds scale novelty (EN) and familiarity (vCA1) recognition might occur though alternating rather than simultaneous activity of brain microcircuits. Taken together, we propose a model for the role of ENvCA1-proj. neurons in recognition memory by balancing memory-guided attentional responses to familiarity and novelty through combination of feedforward inhibition of vCA1 (a node for recognition of familiarity) and the recurrent circuits that contribute to sustain attention to novelty (Fig. 9B).

The model of promoting novelty detection by suppressing familiarity responses is consistent with previous observations showing that vCA1 pyramidal neurons predominantly respond to familiar stimuli whereas novelty response can only occur through activation of vCA1 interneurons (52, 53). Nevertheless, alternating activation of EN and vCA1 could prove essential for the behavioral relevance of EN activity associated with recognition memory, without affecting innate exploratory behavior to novelty. Accordingly, the latter behavior is primarily associated with the prefrontal cortex and amygdala, areas that are only weakly targeted by ENvCA1-proj. neurons (Fig. 2P, S2C) (54–56).

In addition to advancing our understanding of the basic organization and function of brain circuits underlying higher cognitive processes, our findings suggest dysfunction of ENvCA1-proj. neurons could be a key contributing factor to the deficits in CLA complex function and recognition memory found in neuropsychiatric disorders (9, 10). Discrete EN populations can thus emerge as pathophysiological substrate, but also as important treatment targets of key symptoms of these disorders.

Materials and methods

Mice

All experiments were performed in accordance with standard ethical guidelines and were approved by the Danish national animal experiment committee (License number: 2021-15-0201-00801). Unless otherwise noted, all mice used were C57BL6/J mice or transgenic mice with C57BL6/J backgrounds. They were 2-4 months old at the time of the experiment. Transgenic mice used were B6.Cg-Gt (ROSA)26Sortm14(CAG-tdTomato)Hze/J mice (Ai14, JAX007914) and B6J.129S6(FVB)-Slc32a1tm2(cre)Lowl/MwarJ mice (Vgat-Cre, JAX028862). Similar number of male and female mice were used for all experiments except for behavior tests. For behavior tests, male mice were used.

Viruses and retrograde tracers

The information on viruses used in our experiments was the following. CAV-2-Cre (PVM); AAV1-mDlx-HBB-chl-dlox-TVA_2A_oG(rev)-dlox (v271-1, VVF); AAV1-hSyn1-dlox-TVA_2A_mCherry_2A_oG(rev)_dlox (v306-1, VVF); Rabies-GFP (NTNU viral core facility); Rabies-mCherry (NTNU viral core facility); AAV5-CAG-GFP (37525-AAV5, Addgene); AAV5-CAG.hChR2(H134R)-mCherry (10054-AAV5, Addgene); AAV8-hSyn-DIO-hM4Di (Gi) – mCherry (44362-AAV8, Addgene); AAV5-hEF1a-dlox-EGFP(rev)-dlox (v217-5, VVF); and retrograde AAVrg-CAG-GFP (37825-AAVrg, Addgene), AAV1-hSyn1-dlox-jGCaMP8s(rev) (v627-1, VVF). We also used retrograde traces of red Retrobeads (Lumafluor) and Cholera toxin subunit B conjugated with Alexa 647 (CTB647, Thermo Fisher).

Stereotaxic injections

Stereotaxic injections were performed using a stereotaxic frame (Model 940, Kopf). Mice were anesthetized with isoflurane and were subcutaneously injected with buprenorphine (0.1 mg/kg) and metacam (1.5 mg/Kg) for post-operative pain relief. After incising the scalp over the cranium, a small hole was bored with a microdrill bit, and a beveled glass pipette (Wiretrol II, 5-000-2010, Drummond Scientific Company) back-filled with mineral oil and front-filled with the material to be injected was slowly inserted into a target coordinate. After injecting a small volume (50 - 100 nL) with a custom-made displacement injector (based on MO-10, Narishige), the pipette was left in place for 3 to 5 mins before slow retraction. The incision was closed with a nylon suture. The stereotaxic coordinates used were (relative to bregma, in mm): anteroposterior (AP) −3.0; mediolateral (ML) 3.12; dorsoventral (DV) 3.7 for ventral CA1; AP +2.0; ML 0.2; DV 1.5 and 2.5 for prefrontal cortex; AP 0.0; ML 0.2; DV 1.0 and 1.5 for anterior cingulate cortex; AP 0.0; ML 1.5; DV 0.3 and 0.7 for motor cortex; and AP −5.0; ML 3.0; DV 1.5 for dorsomedial entorhinal cortex. For endopiriform injection, we injected two sites from the following coordinates: AP −0.27, ML 3.2, DV 4.2; AP 0.0, ML 3.0, DV 4.2; or AP +1.0, ML 2.7, DV 4.0. Mice were thermally supported with a feedback-controlled heating pad maintained at ∼37 °C (ThermoStar Homeothermic system, RWD). Mice were used for experiments 3-5 weeks post injection.

Implantation of the optic probe

After stereotaxic injection of CAV-2-Cre into the vCA1 and AAV1-hSyn1-dlox-jGCaMP8s into the EN (AP −0.27, ML 3.2, DV 4.2), a fiber optic cannula (400 μm core diameter, 4.5 mm length, NA 0.39, R-FOC-BF-400C-39NA, RWD) was inserted towards EN. Once in the correct depth, exposed brain surface around the implant was covered by Kwik-Cast (WPI) and was then fixed on skull with dental adhesive resin cement (Super-Bond, SUN MEDICAL).

Ex vivo electrophysiology

Mice were deeply anesthetized with isoflurane and decapitated. Horizontal sections (300 µm) containing ventral hippocampus were prepared by vibratome (VT1200S, Leica) in ice-cold choline solutions containing (in mM): 25 NaHCO3, 1.25 NaH2PO4-H2O, 2.5 KCl, 0.5 CaCl2, 7 MgCl2, 25 D-glucose, 110 Choline chloride, 11.6 Ascorbic acid, and 3.1 C3H3NaO3. Slices were subsequently incubated in artificial cerebrospinal fluid (ACSF) containing (in mM): 125 NaCl, 25 NaHCO3, 1.25 NaH2PO4-H2O, 2.5 KCl, 11 D-glucose, 2 CaCl2, and 1 MgCl at 34 °C for 30 mins then at room temperature (∼20 °C) for at least 1 hour before the recording.

Whole-cell recording was performed with an upright microscope (BX51WI, Olympus) equipped with a motor-controlled stage and focus (MP285A and MPC-200, Sutter instrument), differential interference contrast, coolLED (model pE300), and a monochrome camera (Moment, Teledyne). Neurons were visualized with a 60x lens (1.00 NA, LUMPlanFL N, Olympus) with the software Micro-Manager-2.0 gamma (US National Institutes of Health). Pipettes (4-5 MΩ) were pulled from thick-walled borosilicate capillary glass with a puller (model P-1000, Sutter instrument). For a voltage-clamp recording, the internal solution contained (in mM): 135 CsMeSO3, 10 HEPES, 4 Mg-ATP, 0.3 Na-GTP, 8 Na2-Phosphocreatine, 3.3 QX-314 (pH was adjusted to 7.35 with CsOH). In some experiments, a biocytin (4 mg/mL) and Alexa488/586 (50 μM) were further added for morphological studies. Recordings and hardware control were performed with Wavesurfer (Janelia Farm). Signals were amplified and Bessel filtered at 4 kHz with an amplifier (MultiClamp 700B, Molecular Device), then sampled at 10 kHz with a data acquisition board (USB-6343, National Instrument). Recording with access resistance change of > 20% from the baseline (∼30 MΩ or less) were discarded. Liquid junction potential was not corrected. All recordings were performed at ∼34 °C maintained with the inline heating system (TC-324C, Warner instrument).

To stimulated channelrhodopsin-2, a blue LED was delivered at a short pulse (5-msec) through a 4x objective lens (0.16 NA, UPlanSApo, Olympus). The light power (measured at the focal plane) was 20 mW for the recordings in normal ACSF and 40 mW for the recordings in the presence of tetrodotoxin and 4-aminopyridine. Photo-stimulation was repeated 3 to 6 times (at intervals of 10 or 20 sec) to obtain an average response trace. Photo-evoked responses less than 2 standard deviations of the baseline were considered as a “zero” response. The mean amplitudes indicate mean amplitudes between LED stimulus-onset to 50 ms of average responses.

Drugs used for ex vivo electrophysiology were purchased from Merck, TOCRIS, and Hellobio.

Quantification of presynaptic neurons

Mice were deeply anaesthetized with a ketamine (120 mg/kg) and xylazine (24 mg/kg) mixture and were transcardially perfused with 4% paraformaldehyde (PFA). The brains were extracted and immersed in 4% PFA overnight followed by 20 and 30% sucrose solution for a cryoprotection. The brains were subsequently embedded in O.C.T compound (Tissue TEK, Sakura) in a mold and frozen on the dry ice. Coronal slices (100 µm) of entire brain were cut with Cryostat (CM3050S, Leica) and were washed with phosphate-buffered saline (PBS) before DAPI staining. All sections were mounted onto glass slides with a coverslip (thickness:0.13-0.17 mm, Hounisen) and mounting media (DAKO, S3023, Agilent Technologies).

The injection sites and its specificity were verified by registering the images into Allen Brain atlas using QuickNII and VisuAlign (57).

For the visualization and quantification of presynaptic neuron distribution in a 3D mouse brain, epifluorescence images were acquired using a Slidescanner (4231 x 3462 pixels, 10x objective, Olympus VS120) and analyzed with the MATLAB-based program AMaSiNe (34). Specifically, for Fig. 1, coronal sections between AP +3 mm and −4 mm were registered in AMaSiNe, which semi-automatically detects fluorescently labeled presynaptic cells across all sections. To control for transfection variability (which may arise from slight difference in injection volume or site), we normalized the number of presynaptic neurons in the specified brain areas (shown in Fig. 1H) by the total number of presynaptic neurons in those regions.

For the monosynaptic rabies tracing study in Fig. 6, the brain areas containing over 1% presynaptic cells of total presynaptic cells were proceeded further analysis. Fraction of presynaptic cells in the different areas were calculated by dividing a number of presynaptic cells in each brain areas by the average number of presynaptic neurons in all brain areas.

For the study in Fig. 2, the sections were immunostained with NeuN before imaging. Here, sections were immersed with 5% normal goat serum and 0.2% Triton X for 30 mins at room temperature before incubation with NeuN primary antibody (anti-rabbit, 1:1000) (ABN78, Milipore) at 4 °C for overnight. The sections were then washed with PBS and were further incubated with secondary antibody (goat anti-rabbit conjugated with Alexa 488 or 568, 1:500) (A11008 or A11036, Thermo Fisher) for 1 hour at room temperature before mounting. A single plane image of CLA complex was acquired with confocal microscope (10x objective, 1024 x 1024 pixels, Zeiss). 2-4 images, range of AP: +1.4 to −0.2, were collected per brain. Fraction of double-labelled neurons (with different retrograde tracers) were calculated for each brain by dividing a total number of double-labeled neurons by the total number of vCA1-projecting neurons (labeled with one color).

Quantification of fluorescence signal from axons

For Fig. 3 and S3, horizontal section (100 µm) images were taken with Sliderscanner (10x lens, pixel: 2560 x 3072). All images from the same brain were taken with the same light intensity and detection setting. After background subtraction, GFP signals in ventral, intermediate, and dorsal CA1 column were normalized by the highest signal in ventral CA1 in the same brain.

For Fig. 2, coronal sections (100 µm) were immunostained with GFP antibody before imaging. Here, sections were washed with PBS, the sections were incubated in blocking solution containing 5% normal goat serum and 0.2% Triton X for 30 mins at room temperature before incubation with GFP primary antibody (anti-rabbit, 1:1000) (NB600-308, MNovus Biologicals) in blocking solution at 4 °C for overnight. The sections were then washed with PBS and were further incubated with secondary antibody (goat anti-rabbit conjugated with Alexa 488, 1:300) (A11008, Thermo Fisher) for 1 hour at room temperature before mounting. Images were taken with Slidescanner (10x, pixel: 3072 x 3584). Signal per pixel in ROI was calculated for each projection site after background subtraction.

Social memory test with chemogenetic inhibition of ENvCA1-proj. neurons

Social memory test is based on previous articles with slight modifications (58), (59). All tested mice were habituated to the handling for 3 days. The water-soluble CNO purchased from Hellobio was dissolved by saline (final concentration was 0.1mg/ml). Prior to the test day, we divided subject mice into two groups. On the 1st test day (session 1), one group received chlozapine-N-oxide (CNO, 1 mg/kg, i.p.) and another group received a vehicle. After 30 mins, the subject mice were placed in a three-chamber box for 5 mins (habituation), then the same box with two pencil chambers containing the same object in opposite corners for 5 mins (pretest). After 5 mins break in their home cage, the subject mice were again placed in the same box, but this time, an object in one pencil chamber was replaced with different object and an object in another pencil chamber was replaced with a conspecific (3-4 weeks old male, habituated to the handling and pencil chamber for 3 days prior to the session 1). This “sociability test” lasted for 10 mins and was followed by social discrimination test after 30 mins break in the home cage. The social discrimination test lasted for 5 mins and consisted of the same setting as the sociability test except an object in a pencil chamber, which was replaced with a new conspecific (i.e., a novel mice). In the following day, CNO and saline treatments were swapped between the groups, and all mice went through the same series of tests (session 2) for the within-subject comparison of the CNO effect. The position of the pencil chamber including an object or a conspecific were counterbalanced. The tests were performed under 6-10 lx room light and video was recorded at 25 frames/sec with GigE camera (Basler ace acA1300-60gm, Basler) placed at the ceiling.

Trace fear conditioning

After one week of social memory test, subject mice were examined for their tone-associated fear memory and extinction. One group received CNO while another group received vehicle 45 mins before the test. In conditioning day, mice were placed in a box with a metal grid floor and ethanol odor (context A) and after 3 mins were delivered with 20-sec pure tone (2 kHz, 20 dB) followed by 18-sec time gap (trace) and 2-sec footshock (0.5 mA). The exposure to the same tone-trace-shock sequence was repeated 2 more times with inter-trial interval of 2 mins. In the following day, mice were re-exposed to the context A for 3 mins to assess their fear memory to the context. From day 3, mice were placed in a new context, the context B (smooth floor, breach odor), and went through the same protocol as in conditioning day but without shocks. The process was repeated over 5 days to assess the extinction of the fear memory to the tone and trace. All test was performed in dark (but infrared light was on), and behavior was video recorded at 25 frames/sec with infrared GigE camera (Basler ace acA1300-60gm, Basler).

Novel object recognition test

The novel object recognition test is based on the previous article with slight modifications (58). The subject mice were divided into two groups and CNO treatment was done as same as social memory test. 40 mins after CNO/vehicle i.p. injection, the subject mice were placed in an open field box (50 x 50 cm) for 5 mins (habituation). In following familiarization session, two same shape objects are placed in the open field box, and then the subject mice explored for 10 mins. After 1 hour break in their home cage, the subject mice were again placed in the same box, but one object is replaced by a novel object. The subject mice are allowed to explore the box for 5 min in discrimination session.

Social memory test and novel object recognition test with in vivo fiber photometry

Optic probe implanted mice were habituated handling for 3 days prior the test. In test day, the subject mouse attached with patch code were placed in an open field box (50 x 50 cm) for 5 min (habituation). After habituation, the mouse was place in a home cage for resting and waiting following session. For pretest, pencil chambers containing the same objects were placed at the corners of the open field box. Pretest was tested for 5 min. For sociability test, a novel mouse (4 weeks old, male) or a novel object were placed in pencil chambers. Sociability test took 10 min. After sociability test, the subject mouse return to a home cage and rest for 20 min. After 20 min rest, the subject mouse was tested discrimination session for 5 min. In discrimination test, a novel mouse (4 weeks old, male) or a familiar mouse that was used as the stimulus mouse in sociability test were placed in pencil chambers. Both in sociability and discrimination test, positions of pencil chambers were counterbalanced between subject mice. The tests were performed under 6-10 lux room light and video was recorded at 60 frames/sec with GigE camera (Basler ace acA1300-60gm, Basler) placed at the ceiling. Video acquisitions were controlled by Bonsai. Simultaneous start of video recording and calcium signal recording was controlled by Raspberry Pi.

After one week of social memory test, subject mice were examined for novel object recognition test. The procedure, the arenas and objects were used the same as in the novel object recognition test with chemogenetic inhibition of ENvCA1-proj. neurons.

Behavior data acquisition and analysis

Ethovision (version 16 or 17, Noldus) was used for the hardware control and data acquisition and initial quantification for the behavior. Behavioral variables, including movement trajectory, times spent in predefined areas, total distance traveled, travelling velocity, and freezing were extracted and further analyzed/nested with custom functions in the Matlab (R2020b). For social memory test, interaction zones were defined as an area between the edge of pencil chamber and 3 cm away from the edge. When the nose of the subject mouse was in the zone, the time was counted as interaction time. Sociability index was calculated by dividing total interaction time with mouse by total interaction time with mouse and object. The discrimination index was calculated by dividing total interaction time with a novel mouse by total interaction time with a familiar mouse and a novel mouse. Data from mice were excluded if discrimination index was less than 0.5 during saline treatment. For trace fear conditioning test, mice were considered “freezing” if pixel change between frames was less than 0.1 % for more than 1 sec.

Deeplabcut (ver.2.3) was used for tracking of annotated body points of the subject mice connected with patch code for calcium imaging of ENvCA1-proj. neurons during social memory test and novel object recognition test. Nose points were used for further analysis using Matlab (R2020b). To define interaction zone, reference points was set at the center of pencil chamber.

The areas that distance between nose point and reference point was less than 100 pixel for social memory test and 50 pixel for novel object recognition test were dedicated to interaction zone.

Fiber photometry and analysis

Population activity of ENvCA1-proj. neurons were recorded using fiber photometry (Doric Lenses). GCaMP8s was excited with 470 nm light at 20-30 μW measured at the tip (isosbestic point used was 415 nm (15 μW)). Calcium-dependent and independent signals were collected using lock-in mode. Signals were detected at 10x gain. Recorded signals were butterworth filtered (low-pass at 40Hz) and calcium signals (in z-score) were extracted using Doric neuroscience studio V6 (Doric Lenses). Signals were further down sampled to 60 Hz using Matlab (R2020b) signal analyzer to match the behavioral sampling data (60 fps). Z-score > 2.56 (alpha = 0.01) was determined to be a calcium event. For heat map of cumulative time and calcium events, spatial binning was 50 pixels. To make correlation map (Fig. S7 and S8A), correlations between column of cumulative time heat map and cumulative calcium event heat map were tested using Matlab corr function.

Statistical analysis

Data and statistical analysis were performed with MATLAB R2020b (Mathworks). Unless otherwise noted, group data was presented as median, and first and third quartiles are shown for dispersion. Statistics were performed using a non-parametric test (signed-rank test for paired data and rank-sum test for unpaired data). One-way ANOVA (post hoc test: Tukey-Kramer test or Kruskal-Wallis test) were applied for multi-group comparison. Two-way ANOVA were performed for comparison of multi-factors. Pearson correlation coefficient was applied for correlation analysis. Significance was defined as *p < 0.05, **p < 0.01, ***p < 0.001. Outlier was detected by Grubbe test (GraphPad, significant level: 0.05). Sample number were indicated as “n”, animal number were indicated as “N”.